Cell type-specific control of secondary cell wall formation by Musashi-type translational regulators in Arabidopsis

Version 1

Abstract

Deciphering the mechanism of secondary cell wall/SCW formation in plants is key to understanding their development and the molecular basis of biomass recalcitrance. Although transcriptional regulation is essential for SCW formation, little is known about the implication of post-transcriptional mechanisms in this process. Here we report that two bonafide RNA-binding proteins homologous to the animal translational regulator Musashi, MSIL2 and MSIL4, function redundantly to control SCW formation in Arabidopsis. MSIL2/4 interactomes are similar and enriched in proteins involved in mRNA binding and translational regulation. MSIL2/4 mutations alter SCW formation in the fibers, leading to a reduction in lignin deposition, and an increase of 4-O- glucuronoxylan methylation. In accordance, quantitative proteomics of stems reveal an overaccumulation of glucuronoxylan biosynthetic machinery, including GXM3, in the msil2/4 mutant stem. We showed that MSIL4 immunoprecipitates GXM mRNAs, suggesting a novel aspect of SCW regulation, linking translational control to the regulation of SCW biosynthesis genes.

Introduction

As the main constituent of the terrestrial plant biomass, lignocellulose represents a huge reservoir of fixed carbon and a renewable resource for bioproducts and energy that is essential for human usages1. Lignocellulose biomass enclosed mainly within secondary cell wall (SCW) reinforces vessels for long-distance transport and provides physical properties to fibers allowing upright growth of the plants1, 2. SCWs are composite material made of high molecular weight biopolymers, including cellulose, hemicelluloses, and lignin, as well as cell wall proteins whose composition and structure can differ markedly between dicots and monocots2, 3. Cellulose is synthesized at the plasma membrane by the cellulose synthase complex and is composed of linear chains of β-(1–4) glucans that aggregate to form highly crystalline cellulose microfibrils2, 3. Xylans, the main hemicellulosic polysaccharides in SCW, are synthesized in the Golgi apparatus by a complex biosynthetic machinery and comprise a group of polysaccharides that share a common backbone of β-(1,4)-linked xylose (Xyl) units, but differs by the presence of side chains, whose nature is dependent on tissues and species2. In particular, glucuronoxylan, the dominant xylan in the SCW of dicots, is notably substituted at O-2 position with α-D-glucuronic acid (GlcA) or (4-O-methyl)-α-D-glucuronic acid (MeGlcA) groups, and with O-acetyl moieties at C-2 or C-3 positions2. By interacting with the cellulose and lignin, xylans contribute to the strengthening of the SCW, and are considered to be one of the main factors contributing to the biomass recalcitrance to enzymatic hydrolysis3, 4. Finally, lignin polymers are made up of H, G, and S units which are the results of the oxidation and the polymerisation of the p-coumaryl, coniferyl and synapyl alcohols, respectively. These monomers, also called monolignols are synthesized on the ER surface/cytoplasm prior transportation to the cell wall. Lignins confer stiffness, strength and hydrophobicity to the SCW2–4.

The metabolic load imposed by the coordinated production of SCW biopolymers emphasizes the importance for plants of having a precise control over temporal and spatial expression of the corresponding SCW biosynthetic genes. Based on founding studies performed in Arabidopsis model, it is generally recognized that the SCW synthesis is primarily controlled at a transcriptional level by a multilayer and interconnected network of transcription factors/TFs5, 6. Key players of this regulatory network are related NAC (NAM-ATAF1/2-CUC) and MYB-type TFs that induce, in a redundant and combinatorial manner, the expression of SCW biosynthetic genes by binding their promoter region5. In addition, a family of Class III homeodomain leucine zipper (HD-ZIPIII) TFs, whose members control several aspects of Arabidopsis development, was also shown to interact with SCW regulatory network, contributing to xylem cell specification and SCW synthesis5, 6. Although this transcriptional level of control can gain complexity in perennial plants, functional orthologs of the main regulators were identified in many different species, suggesting that SCW deposition proceeds via a conserved mechanism in all vascular plants7, 8. During the last decade, engineering the SCW regulatory network to deposit modified lignin and/or improve polysaccharide composition became a promising strategy to optimize biomass processability9, 10. However, despite huge progresses in understanding the regulation of SCW synthesis, the engineering of SCW regulatory network to improve biomass yield and digestibility has proven difficult and often impairs plant growth because of altered vascular tissue development11, 12.

Downstream of transcriptional regulations by TFs, RNA-binding proteins (RBPs) are essential post-transcriptional modulators of gene expression across all kingdoms of life13. However, although a large cohort of RBPs exists in plants14, there is little information about RBPs-mediated posttranscriptional regulation of SCW biosynthetic genes. So far, only few microRNAs (miRNAs) families have been implicated in the post-transcriptional regulation of genes involved in both regulatory and enzymatic aspects of SCW biosynthesis6. In particular, miR165/166 were shown to target HD-ZIPIII transcriptional regulators, having a direct impact on the transcriptional network associated with SCW biosynthesis6. In the other hand, miRNA397/857/408 were shown to affect lignin content via the targeting of laccase/LAC genes, which control lignin polymerization from monolignols precursors15, 16. In addition, two related tandem CCCH zinc-finger proteins, which exhibit both DNA and RNA binding abilities, have been proposed to modulate SCW formation by regulating the expression of genes associated with cell wall metabolism17. Despite these mounting evidences, our current knowledge of the role of post-transcriptional regulators in SCW synthesis is still very limited.

In this work, we report that two RRM-domain containing RNA-binding proteins homologous to the animal translational regulator Musashi, Musashi-like2/MSIL2 and Musashi-like4/MSIL4, function redundantly to control various aspects development, including the stiffness of the inflorescence stem in Arabidopsis. We show that RRM-dependent RNA-binding activity is essential for MSIL2/4 functions in vivo, and that MSIL2/4 interactomes are similar, being enriched in proteins involved in 3’-UTR binding and translational regulation. MSIL2/4 mutations alter the formation of SCW in the interfascicular fiber cells, leading to a reduction in lignin deposition, and a change in the decoration pattern of glucuronoxylan that is associated with an increase of 4-O-methylation of GlcA substituent. In accordance, quantitative mass-spectrometry-based protein analysis reveal an overaccumulation of glucuronoxylan biosynthetic machinery, including the GlucuronoXylan Methyltransferase3/GXM3, in the msil2/4 mutant stem. We show that MSIL4 immunoprecipitates GXM3 mRNA in vivo, likely regulating its expression at a translational level. Our results demonstrate that MSILs regulate SCW synthesis in interfascicular fiber cells and point to a novel aspect of SCW regulation linking translational repression to regulation of SCW biosynthesis genes.

Results

Musashi-like MSIL2/4 proteins redundantly control specific Arabidopsis development processes

While many studies have implicated the MSIs in the control of gene expression during cellular proliferation, cell fate determination and cancer in animals18, 19, yet little is known about the presence and activity of MSI-type proteins in plants. In this work, we describe an RNA-binding protein family in Arabidopsis, whose members hereafter named as MUSASHI- Like1 to 4/MSIL1-4, share both sequence, domain organization, and model-based structural similarities with animal MSIs (Fig. 1a, b). Phylogenetic analysis revealed that Arabidopsis indeed harbors a clade of seven MSIL-type genes with a prominent sub-clade harboring the MSIL1-4 genes (Fig. 1b). Notably, a neighboring sub-clade contains two genes, RBGD2 and RGBD4 (Fig. 1b), that encode for heat-inducible RBPs implicated in the response to heat stress in Arabidopsis20. MSIL1-4 proteins share a common domain architecture consisting of two N-terminal RNA recognition motifs (RRM1 and RRM2) that are followed by a poorly conserved, intrinsically unstructured, carboxy-terminal region (Fig. 1a; and Supplementary Fig. 1a). The survey of the gene expression ARAPORT11 databases (https://araport.org/; http://fgcz-pep2pro.uzh.ch) revealed that the MSIL1-4 genes are widely expressed in Arabidopsis, with MSIL1 showing the lowest levels of expression (Fig. 1c). Database searches further indicated that MSIL orthologs are widely distributed in land plants, from bryophytes to angiosperms (Supplementary Fig. 1b). To gain a mechanistic understanding of MSIL function in Arabidopsis, we determined the subcellular localization of these proteins by performing fluorescence microscopy on GFP-tagged versions of MSIL2 and MSIL4 (MSIL2G and MSIL4G). Both proteins showed a diffuse cytoplasmic distribution in root cells of stable Arabidopsis transgenic plants that contrasts with the nucleus and cytoplasmic distributions of the free GFP protein control (Fig. 1d).

Musashi-like MSIL2/4 proteins redundantly control development in Arabidopsis.

a) Schematic representation of the Musashi-like (MSIL) protein family in Arabidopsis thaliana. The invariant RNP1 and RNP2 motifs within the conserved RRM domains are indicated. Numbers refer to amino acid identities between the Mouse Musashi/MSI RRM1/2 domains and the corresponding domains in Arabidopsis MSIL homologs. b) Evolutionary relationships between MSILs and related RNA-binding proteins. The scale bar indicates the rate of evolutionary change expressed as number of amino acid substitutions per site. c) RNA- seq expression map of MSIL genes extracted from ARAPORT11. Shading is a log2 scale of transcripts per million (TPM). d) General overview of the roots of 6-days-old wild type Arabidopsis seedlings (Col-0) or transgenic seedlings expressing either a free GFP protein or GFP-tagged versions of MSIL2/MSIL2G and MSIL4/MSIL4G. Scale bar, 10 μm. e) Photographs of representative rosettes of Col-0, msil2/4, msil2/4-MSIL2F1 and msil2/4- MSIL4F1 plants. f) Photographs of representative inflorescence stems of Col-0, msil2/4, msil2/4-MSIL2F1 and msil2/4-MSIL4F1 plants. Abbreviations: Mus musculus Musashi2 (MmMsi2); human heterogeneous nuclear ribonucleoprotein A1 (hnRNPA1).

To further investigate the function of the MSILs in Arabidopsis, we raised specific antibodies against non-conserved regions of MSIL1-4 and characterized T-DNA insertion msil mutant lines that lack full-length MSIL mRNA and MSIL proteins as judged by RT-PCR and western blot experiments, respectively (Supplementary Fig. 2a-c). However, none of the four characterized single msil mutants showed any obvious phenotype (Supplementary Fig. 3a and data not shown). To assess the potential functional redundancy between MSIL genes, we crossed the single msil mutants together and characterized the six possible double mutant combinations (Supplementary Fig. 3a). Only the msil2-1/msil4-1 (msil2/4) double mutant showed discernable developmental abnormal phenotypes, including enlarged and curled leaves, early leaf senescence and a pendant inflorescence stem (Fig. 1e, f; and Supplementary Fig. 3a, b). To assess whether MSIL1 and MSIL3 also contribute to Arabidopsis development in the msil2/4 double mutant background, we generated a msil1-1/msil2-1/msil3-1/msil4-1 (msil1/2/3/4) quadruple mutant that shows no particular or aggravated developmental phenotypes with respect to the msil2/4 double mutant (Supplementary Fig. 3a, c, d). To test the functional redundancy of MSIL2 and MSIL4 proteins in vivo, we transformed the msil2/4 double mutant with constructs coding either for a Flag/HA-tagged version of MSIL2 (MSIL2F) or MSIL4 (MSIL4F), and selected two independent transgenic plants expressing similar and near physiological levels of the tagged protein, as evaluated by western blot using our home-made antibodies (Supplementary Fig. 3e). The ectopic expression of either MSIL2F or MSIL4F proteins was able to rescue all the phenotypes associated with the MSIL2/4 defect (Fig. 1e, f; and Supplementary Fig. 3b). This confirms that MSIL2F and MSIL4F are functional proteins and more importantly that MSIL2 and MSIL4 redundantly control various aspects of development in Arabidopsis.

RRM-dependent RNA-binding activity is essential for MSIL2/4 functions in vivo

Animal MSIs interact with target transcripts through their RRM domains and mutations in these domains have been shown to impair MSI activity in vivo21, 22. MSILs are very similar to metazoan MSIs over their RRM sequences (Fig. 1a; and Supplementary Fig. 4a), a closeness of sequences also observed when homology models of the MSIL4 RRM motifs were generated using SWISS-MODEL server23 (Fig. 2a). Homology-based structure prediction proposed that these domains adopt the characteristic RRM fold with a four-stranded β-sheet structure bearing the conserved phenylalanine residues involved in the specific recognition of RNA bases24 (Fig. 2a; and Supplementary Fig. 4a). This supports the idea that MSILs have retained the ability to interact with RNA, a notion further supported by the identification of MSIL2/4 proteins in the experimentally determined Arabidopsis mRNA-binding proteome14 To examine the requirement for the interaction of MSIL with RNA in developmental control, we first generated an RNA-binding mutant form of MSIL4, MSIL4RRM, by introducing phenylalanine to aspartate (F→D) mutations in both RRM domains (Fig. 2b). Protein-RNA interaction assays using RNA homopolymers immobilized on agarose beads confirmed that the MSIL4, but not MSIL4RRM, exhibited intrinsic RNA binding activity in vitro (Fig. 2c). To assess the impact of these mutations in vivo, the MSIL4G and MSIL4GRRM constructs were expressed under the control of the MSIL4 endogenous promoter into the msil2/4 mutant, and two independent transformants expressing near physiological levels of the WT (MSIL4G-3 and MSIL4G-8), and mutant (MSIL4GRRM-3 and MSIL4GRRM-10) proteins were selected for further complementation analysis (Supplementary Fig. 4b). MSIL4G, but not MSIL4GRRM proteins were able to rescue the developmental defects incurred by MSIL2/4 mutations (Fig. 2d; and Supplementary Fig. 5a, b), indicating that these proteins are likely to exert their functions by interacting with RNA targets in vivo.

RRM-dependent RNA-binding activity is essential for MSIL function in planta.

Experimentally determined structures of the RRM1 domain of MmMSI2 (PDB ID: 6C8U) and the RRM2 domain of MmMSI1 (PDB ID: 5x3y), and models of the RRM domains of MSIL4 generated using the homology-modeling server SWISS-MODEL. The α-helices and β-sheets of RRM domain as well as the phenylalanine residues that contact RNA bases are highlighted. b) Schematic representation of the mutations introduced in the RRM domains of MSIL4. c) Binding assays of MSIL4 RRM domains on ssRNA homopolymers in vitro. Single-stranded polyA (pA), polyG (pG), polyC (pC) or polyU (pU) RNAs were subjected to binding with the His-tagged recombinant RRM domains from MSIL4 (upper part) or its mutated version (MSIL4RRM, lower part). An anti-His antibody was used for detection. In. represents the input fraction. d) RRM-dependent RNA-binding activity is essential for MSIL2/4 function in stem. Photographs of representative inflorescence stems of Col-0 and msil2/4 mutants complemented or not with a WT version of MSIL4 (MSIL4G-8 and 3) or an rrm mutant (MSIL4GRRM-3 and 10).

MSIL2/4 protein interactomes are enriched in proteins involved in 3’-UTR binding and translational regulation

Animal MSIs modulate mRNA expression mostly at a translational level through binding to conserved motifs in the 3’-UTR of the target mRNA and further interactions with various mRNA binding partners18, 25, 26. To further characterize the components of the MSIL2/4 network in Arabidopsis, we performed affinity purification coupled to LC-MS/MS, as described in Scheer et al.27, using the complemented MSIL2F1 and MSIL4F1 lines. Consistent with their functional redundancy, MSIL2F1 and MSIL4F1 exhibit a similar protein interaction network, with 27 proteins significantly enriched in both MS analyses (Fig. 3a, b; and Supplementary Table 1). Examination of the PANTHER database (http://www.pantherdb.org) identified that MSIL2F1/4F1 are part of a protein-protein interaction network that was principally enriched in 4 categories of GO molecular functions, namely poly(A) binding protein (GO:0008143), mRNA 3-UTR binding (GO:0003730), single-stranded RNA binding (GO:0003727), and translation initiation factor activity (GO:0003743), highlighting the relationship between MSIL2/4 and the mRNA binding and regulatory proteome (Fig. 3c).

MSIL2/4 protein interactomes are enriched in proteins involved in 3’-UTR binding and translation regulation.

a) The semi-volcano plots show the enrichment of proteins co-purified with MSIL2F1 and MSIL4F1 as compared with control IPs. Y- and X-axis display adjusted p-values and fold changes, respectively. The dashed lines indicate the threshold above which proteins are significantly enriched/depleted (fold change > 2; adjP < 0.05). b) Venn diagram showing the overlap between the MSIL2 and the MSIL4 interactomes. c) PANTHER-based classification of the GO molecular function that are overrepresented among the MSIL-interacting proteins. d) Confocal monitoring of the colocalization of MSIL2G, MSIL4G and PAB2R in root tips of 8-d-old seedlings were monitored after 30 min of exposure to 38°C (heat stress) or 20°C for control treatment. e) Cycloheximide treatment inhibits the formation of MSILG and PAB2R cytosolic foci. The transgenic plants expressing the stress granule marker PAB2R was used as a control. Root of 7-day-old seedlings expressing either MSIL4G, MSIL2G, or PAB2R were monitored after 1 hour of exposure to 38°C in the presence of DMSO control treatment (DMSO) or in the presence of cycloheximide inhibitor (+CHX). Scale bar, 10 μm.

Many of the proteins present in the MSIL2/4 interactome, such as PABs, PUMs, UBP1 are involved in translation regulation and are targeted to stress granule (SGs) components, a membraneless cytoplasmic organelle that formed consecutive to the global inhibition of translation caused by heat-stress28, 29. To assess the potential functional link between MSIL and those proteins, we crossed the MSIL2G/4G plants with a functional PAB2-RFP/PAB2R line29, and assessed the localization of the fusion proteins before and after heat-stress in both parent and crossed lines. Fluorescence analysis revealed that MSIL2G/4G, like PAB2R, show a dispersed cytoplasmic signal under normal condition, and are recruited to cytoplasmic granules that overlap with PAB2R-containing SG upon heat stress (Fig. 3d; and Supplementary Fig. 6). Treatment with cycloheximide (CHX), a translation inhibitor that prevent SG assembly by trapping mRNAs at polysomes30, confirmed that those foci correspond to bonafide SGs (Fig. 3e). These results were consistent with the RNA-binding protein-based signature and the nature of MSIL interactomes, further supporting a functional relationship between MSILs and the translation machinery in Arabidopsis.

MSIL2/4 proteins regulate the molecular architecture of SCW in Arabidopsis fibers

Pendant stem inflorescence is a common phenotypic trait of Arabidopsis mutants defective in SCW biosynthesis, that results from a lack of mechanical support and rigidity6, 31. Given the current lack of knowledge of the role of post-transcriptional mechanisms in the control of SCW biosynthesis in plants, we henceforth focused our analysis on the pendant stem phenotype of the msil2/4 double mutant. Survey of the Arabidopsis inflorescence stem tissue-specific transcriptome database32 indicated that the MSIL2/4 genes are highly expressed in the SCW-forming xylary fiber (F) and xylem vessel (X) cells (Fig. 4a). Toluidine blue O staining and microscopic analysis of Col-0 and msil2/4 mutant stem cross-sections revealed no major changes in the architecture of both the interfascicular and xylary fibers in the mutant stem. Notably, xylem cells deformation or collapse, which are a characteristic of mutants deficient in SCW synthesis, were not observed in the msil2/4 mutant (Fig. 4b). However, the examination of the cell wall thickness in cross sections of stems revealed that the interfascicular fiber cells display a significantly thinner SCW in msil2/4 mutant (Fig. 4c). UV autofluorescence and phloroglucinol staining of the Col-0, msil2/4, msil2/4+MSIL2F1 and msil2/4+MSIL4F1 stem sections revealed a reduced lignin signal in the interfascicular fibers of the msil2/4 mutant compared to Col-0 or complemented plants (Fig. 4d; and supplementary Fig. 7a). In agreement with these observations, a significant decrease of 30% in AcBr lignin content was observed in the msil2/4 mutant compared to the Col-0 control plants (Fig. 4e). Despite a tendency to decrease, the difference of lignin content was not significant in the complemented lines msil2/4+MSIL2F1 and msil2/4+MSIL4F1 compared to Col-0 (Fig. 4e). To further investigate potential changes in the lignin structure and composition in the Col-0 and msil2/4 mutant, we evaluated the impact of MSIL2/4 deficiency on lignin composition by thioacidolysis, a degradation method allowing the determination of the relative amounts of H, G and S lignin units linked by beta-O-4 linkages. Total monomer yield did not significantly differ between the col-0 and the msil2/4 mutant (Supplementary Fig. 7b), suggesting that the structure of the lignin polymer is not merely affected in the mutant. Further examination of β-glucan composition using the fluorescent dye calcofluor white33 revealed an increased staining in the interfascicular fiber cells of msil2/4 mutant, supporting changes in the composition/accessibility of cellulose (Fig. 4f; and Supplementary Fig. 7c).

MSIL proteins regulate the molecular architecture of SCW in Arabidopsis fibers.

a) Expression profiles of the MSIL2 and MSIL4 genes in the inflorescence stem tissues as retrieved from the Arabidopsis inflorescence stem tissue-specific transcriptome database40 (https://arabidopsis-stem.cos.uni-heidelberg.de/). The results of two replicates are shown. F, fibers; X, xylem vessels; Cx Cambium (xylem side); Cp, Cambium (Phloem side); P, Phloem; S, Starch sheath; E, Epidermis. b) Cross-sections of Col-0 and msil2/4 stems showing vascular bundle and interfascicular fiber cells stained with Toluidine blue. c) Measurements of the SCW thickness of xylary fiber and interfascicular fiber cells. Values are means (n>400)± SEM. Data were analyzed by unpaired Student’s test. Asterisks indicate significant differences relative to Col-0; ****p<0.0001. d) Top: lignin autofluorescence under ultraviolet (UV) using confocal microscopy in Col-0, msil2/4 mutant, and complemented msil2/4- MSIL2F1 and msil2/4-MSIL4F1 plant stem sections. xy: xylem fibers; if: interfascicular fibers. Bottom: Phloroglucinol-HCl staining of Col-0, msil2/4, and complemented msil2/4-MSIL2F1 and msil2/4-MSIL4F1 plant stem sections. Scale bar, 100 μm. e) Histograms represent acetyl bromide lignin content, expressed as % of dry weight. Lignin content was measured in the bottom section of mature stems (3 biological replicates per line). Significant differences between Col-0, msil2/4+MSIL2F1, msil2/4+MSIL4F1, and msil2/4 plant lines are indicated by asterisks*, P-value< 0.01 according to Student’s t test (n = 6 to 10). f) Confocal microscopy of Calcofluor-white staining of cross sections of Col-0 and msil2/4 mutant inflorescence stems. Scale bar, 100 μm. g) Glucose yield after incubation of Col-0 and msil2/4 lignocellulosic material with cellulases either without pre-treatment (-Tr) or after sodium hydroxide pre-treatment (+Tr).

To extend this analysis, we then subjected purified Col-0 and msil2/4 SCW materials to Fourier-Transform Infrared (FTIR) spectroscopy34. PLS-DA analyses performed on FT-IR absorbance data clearly separated msil2/4 mutant lines from Col-0 plants, using the first two components (explaining 74% and 10% of total variability, respectively) (Supplementary Fig. 7d, left panel). The large-scale chemotyping of CW composition in stem by FTIR spectroscopy suggested that not only lignin deposition was impaired in msil2/4 mutants but also polysaccharides composition (Supplementary Fig. 7d, right panel). The lowest absorbance values in msil2/4 lines were observed in the region 1800-1500cm-1, mostly associated to the absorbance of lignin related compounds. On the contrary, a higher absorbance was detected in msil2/4 samples within 1050-900 cm-1 region, mostly associated to polysaccharides compounds34. Because the optimal interaction between the SCW lignin and polysaccharide polymers is believed to contribute to SCW recalcitrance, we used saccharification as a proxy to compare cell wall sugars accessibility between Col-0, msil2/4 and complemented mutant lines (Fig. 4g). A higher amount of reducing sugars from native cell wall (without pre-treatment/-Tr) was released using msil2/4 samples compared to Col-0 samples. After cell wall loosening using alkali pre-treatment/+Tr, an even more important amounts of sugars (+5%) was released from double mutants CW enzymatic digestion (Fig. 4g). Together, our data indicate that MSIL2/4 regulate the molecular architecture of SCW in the interfascicular fibers, contributing significantly to the setting of biomass recalcitrance in Arabidopsis.

The accumulation and activity of the glucuronoxylan decoration machinery are mis-regulated in msil2/4 mutant

To elucidate the impact of the MSIL2/4 on the expression of SCW-related biosynthesis genes in the Arabidopsis inflorescence stem, we performed mRNA sequencing (mRNA-seq) on both wild-type and msil2/4 mutant backgrounds. In total, we found 234 genes to be significantly differentially regulated (DEG) in the msil2/4 mutant background (with stringent thresholds having a logarithm of fold change >1.5 or <-1.5 and false discovery rate (FDR ≤ 0.05), including 156 up-regulated and 78 down-regulated mRNAs (Supplementary Fig. 8a; and Supplementary Table 2). Interestingly, genes encoding regulatory or enzymatic components of the SCW biosynthesis pathway were absent from the DEG list, and their expression was not significant changed in msil2/4 mutant (Supplementary Fig. 8a). Gene Ontology analysis of DEGs revealed significantly enriched signaling pathways (FDR ≤0.05) involved in plant defense or responses to biotic and abiotic stresses (Supplementary Fig. 8b), encoding products that are preferentially targeted to the extracellular region and the cell wall (Supplementary Fig. 8c). In this regard, the constitutive activation of defense pathways has been well documented in mutant plants defective in cell wall formation, including the SCW35.

The absence of SCW biosynthesis genes in the DEG list suggests that MSIL2/4 could regulate the SCW formation at a post-transcriptional/translational level, a hallmark of MSI function in animal models18, 19. To investigate proteome-level changes that could potentially occur upon MSIL2/4 mutations, we performed a comparative proteomic analysis of Col-0 and msil2/4 inflorescence stems using MS-based quantitative proteomics, and represented ≈4300 reliably identified and quantified proteins in a volcano plot according to their statistical p-value and their relative difference of abundance. This analysis revealed 267 proteins with significant abundances in the msil2/4 mutant versus Col-0 inflorescence stems (fold change ≥2, Benjamini-Hochberg FDR < 1%), including 156 up-regulated, and 111 down-regulated proteins (Fig. 5a; and Supplementary Table 3). While no specific GO terms were enriched in the set of down-regulated proteins, the inspection of the up-regulated proteins revealed a strong enrichment in enzymes involved in the glucuronoxylan biosynthetic pathway (Fig. 5a, b; and supplementary Table 3). Notably, the upregulated glucuronoxylan-related proteins included glycosyltransferases required for the xylan backbone synthesis (IRX9/10/14), as well as enzymes involved in its substitution (ESK1, GUX1, and GXM3)2 (Fig. 5a, c). Visual inspection of the RNA-seq aligned reads of the up-regulated glucuronoxylan-related genes using the Integrative Genomics Viewer (IGV) and quantitative RT-PCR analysis of the glucuronoxylan-related decoration genes confirmed that the observed proteomic changes in msil2/4 were not associated with significant variations in mRNA levels (Supplementary Fig. 9a, b).

The accumulation and activity of the glucuronoxylan decoration machinery are altered in msil2/4 mutant.

a) MS-based quantitative comparison of Col-0 and msil2/4 inflorescence stem proteomes. Volcano plot displaying the differential abundance of proteins in Col-0 and msil2/4 cells analyzed by MS-based label-free quantitative proteomics. The volcano plot represents the - log10 (limma p-value) on y axis plotted against the log2 (Fold Change msil2/4 vs Col-0) on x axis for each quantified protein. Green and red dots represent proteins found significantly enriched respectively in msil2/4 and Col-0 Arabidopsis stems (log2(Fold Change) ≥ 1 and - log10(p-value) ≥ 2.29, leading to a Benjamini-Hochberg FDR = 1.01 %). The up-regulated proteins involved the glucuronoxylan biosynthetic pathways are indicated. b) Gene Ontology (GO) analysis of the proteins that are significantly up-regulated in the msil2/4 mutant. The gene ontology analysis of DEGs was performed using ShinyGO v0.76 software. Lollipop diagrams provide information about GO fold enrichment, significance (FDR in log10), and number of genes in each pathway. c) Schematic model of glucuronoxylan substitution patterns in Arabidopsis and the enzymes involved. IRX9/10/14, glycosyltransferases involved in the synthesis of xylan (Xy) backbone. ESK1, eskimo1. GUX1, glucuronic acid substitution of xylan1; GXM3, glucuronoxylan methyltransferase3 are involved in the glucuronoxylan decoration. d) Neutral monosaccharide composition of alcohol insoluble residue (AIR) extracted from inflorescence stems of wild-type and msil2/4 mutant plants that have been pre- hydrolyzed or not with acid. Rha, rhamnose; Fuc, fucose; Ara, arabinose; Xyl, xylose; Man, mannose; Gal, galactose; Glc, glucose. e) Left: Immunofluorescence labeling of transverse sections of Col-0 and msil2/4 stems with the LM10 and LM11 antixylan antibodies. Xy, xylem fibers; if, interfascicular fibers. Right: Quantification of the fluorescence was done using ImageJ software and processed according to Supplementary table 4.

The outcomes of our histological and proteomic analyses suggest potential variations in polysaccharide content and/or composition in the SCW of msil2/4 mutant. To address this point, we first performed glycome profiling on either untreated (-Tr) or H2SO4-treated (+Tr) stem samples. The levels of glucose (Glc) and xylose (Xyl), which are the building blocks of cellulose and xylan polysaccharides in the SCW were not substantially affected in the msil2/4 mutant (Fig. 5d). Similarly, no significant changes were observed in the levels of the primary cell wall-related monosaccharides, galactose, rhamnose, arabinose, mannose, fucose and galacturonic acid (Fig. 5d), indicating that the MSIL2/4 mutations did not significantly altered cellulose or glucuronoxylan levels in the SCW.

To assess possible changes in glucuronoxylan decoration in msil2/4 mutant, we performed cytochemical analysis on Col-0 and msil2/4 stem sections using two antixylan antibodies (LM10 and LM11), whose binding specificities depend on the degree of glucuronoxylan substitution36. LM11, which binds with similar affinity both unsubstituted and substituted glucuronoxylan, shows a similar signal in the interfascicular and xylary fibers of Col-0 and msil2/4 mutant stems, confirming that the glucuronoxylan content was not significantly modified in msil2/4 compared to Col-0 (Fig. 5e). In contrast, LM10, which binds less efficiently the substituted glucuronoxylan36, shows a significant signal decrease in the interfascicular fibers of msil2/4 mutant (Fig. 5e; and Supplementary Table 4). Together, our data indicate that the MSIL2/4 mutations impact the accumulation and activity of the glucuronoxylan decoration machinery in the interfascicular fiber cells of Arabidopsis stem.

MSILs restrain the degree of 4-O-methylation of glucuronoxylan in Arabidopsis

In Arabidopsis, the glucuronoxylan backbone is decorated predominantly by acetyl, GlcA, and MeGlcA groups, that are deposited in a regular and controlled manner by a specific enzymatic machinery2, 37, 38. Interestingly, the ESK1, GUX2, and GXM3 enzymes, which are involved in the deposition of the acetyl and GlcA groups, as well as the methylation of GlcA group, respectively, are over-accumulated in the msil2/4 stem proteomics, providing a possible explanation for the modification in glucuronoxylan decoration observed in the interfascicular fiber cells of the msil2/4 mutant. To investigate further the variations in glucuronoxylan decoration incurred by the msil2/4 mutation, we analyzed the patterns of xylanase-released SCW oligosaccharide using matrix-assisted laser desorption ionization-time-of-flight mass spectrometry (MALDI-TOF)39. Consistent with previously observations40, 41, MALDI-TOF analysis of Col-0 glucurononoxylan oligosaccharides showed the release of xylo-oligosaccharide pairs, evenly spaced by their degrees of polymerization and acetylation, bearing a GlcA (m/z 1705.5/1747.5) or methylated GlcA (m/z 1719.5/1761.5) substitution group (Fig. 6a). Interestingly, the level of GlcA branched xylo-oligomers was strongly decreased in the msil2/4 double mutant, and this whatever the complexity of the oligomers considered (Fig. 6a; and Supplementary Fig. 10a). Importantly, the molecular phenotypes observed in the msil2/4 mutant were rescued in both msil2/4+MSIL2F1 and msil2/4+MSIL4F1 complemented lines, confirming the specificity of the mutation (Fig. 6a). In agreement with the MALDI–TOF outcome, quantification of the acidic sugar levels in untreated (-Tr), or H2SO4-treated (+Tr) WT and msil2/4 stem samples confirmed that the release of GlcA moiety was strongly reduced in msil2/4 double mutant, as opposed to the galacturonic acid/GalA group located on pectins (Fig. 6b).

MSILs restrain the degree of 4-O-methylation of glucuroxylan in Arabidopsis inflorescence stem.

a) MALDI-TOF mass spectra of xylooligosaccharides generated by xylanase digestion of xylan from Col-0, msil2/4, and complemented msil2/4-MSIL2F1 and msil2/4-MSIL4F1 inflorescence stem materials. The ions at m/z 1705/1747 and 1719/1761 correspond to acetylated xylo-decapolysaccharides bearing a GlcA residu (GlcA-Xyl10- Ac4//GlcA-Xyl10-Ac5) or a methylated GlcA residue (MeGlcA-Xyl10-Ac4//MeGlca-Xyl10- Ac5). b) Acidic monosaccharide composition of alcohol insoluble residue (AIR) extracted from inflorescence stems of wild-type and msil2/4 mutant plants that have been pre- hydrolyzed or not with acid. GalA, galacturonic acid; GlcA, glucuronic acid. c) Relative changes in unmethylated/methylated GlcA decapolysaccharide ratio in Col-0 (black) and msil2/4 mutant (red) as controlled by the addition of external spike-in control corresponding to a pentaacetyl-chitopentaose. d) GFP-based RNA-IP assays. Top panel (WB): western blots performed using antiGFP or antiUGPase antibodies on protein fractions from inputs or antiGFP immunoprecipitates from WT and the msil2/4 mutant plants expressing the WT (MSIL4G-3) or RRM mutant (MSIL4GRRM-3) MSIL4-GFP fusions. Bottom panel (RT) : Corresponding RT-PCR using GXM1 or GXM3 specific primers on RNA fractions.

It has been previously proposed that the intermediate level of glucuronoxylan methylation observed in Arabidopsis is due to an intrinsic rate-limiting activity of GXM enzymes in vivo42. We noticed that in addition to GXM3, a second member of the GXM family, GXM142, tends also to be over-expressed in msil2/4 stem although at a fold change of 1.77 slightly below the fixed significance threshold in the proteomic analysis (Fig. 5a; and Supplementary Table 3). This observations make us wonder whether the apparent decrease of GlcA branched xylooligosaccharides in msil2/4 could be due to an overmethylation of the GlcA substituent. To address this question, we performed the xylanase digestion and MALDI-TOF analysis in presence of an added spike-in oligosaccharide control (Pentaacetyl- chitopentaose) for normalization of the MALDI_TOF data (Supplementary Fig. 10a). Upon normalization, the MALDI-TOF data indicated that the decrease of the peaks corresponding to the GlcA-branched xylooligosaccharides was correlated with a increase in corresponding meGlcA branched xylooligosaccharides, (Fig. 6c; and Supplementary Fig. 10b; and Supplementary Table 5), suggesting that the MSIL2/4 mutations unleash the activity of the GXM1/3 in stem. RNA immunoprecipitation assays further confirmed that the GXM3 and GXM1 mRNAs are efficiently pulled-down from MSIL4G-3, but not MSIL4GRRM-3 plant extracts (Fig. 6d), suggesting that the MSIL2/4 interaction network controls the 4-O-methylation of glucuronoxylan by repressing the activity of the GXM1/3 mRNAs in stem.

Discussion

RNA-binding proteins (RBPs) are essential components of the gene regulation machinery that govern the fate and expression of cellular RNA at post-transcriptional levels, including processing, splicing, base modification, translation and degradation13. Sequence-based bioinformatic, reverse and forward genetic, and affinity-based proteomic analyses have converged to show that plants, like animals and fungi, harbor a large diversity of RBPs whose functions in gene regulation and plant development are ill-defined14. In this study, we have identified an RRM motif-containing RBP family, hereafter named as MSIL, whose members share sequence, structural and functional similarities with the animal translational regulator Musashi/MSI. We found that two of the four MSIL members, MSIL2 and MSIL4, function redundantly to control various aspects of Arabidopsis development, including leaf senescence, morphology of rosette leaves, and the rigidity of the inflorescence stem. We provide evidence that the RRM domains of the MSIL2/4 proteins exhibit RNA-binding ability and are essential for the activity of these proteins in vivo, an observation that is corroborated by the presence of the MSIL2/4 proteins in an experimentally determined Arabidopsis mRNA-binding proteome14. Further supporting a role for MSIL2/4 in mRNA-related transactions, affinity purification-mass spectrometry analysis revealed that these proteins share a common network of protein-protein interactions with well-known mRNA-binding proteins, including the PAB2/4/8, RBP45/47, and the translation factor EIF4G. Notably, the animal polyA-binding protein/PABP (homolog of the plant PABs) was previously shown to be a functional partner of MSI that is essential for it activity in translation repression18, 26. Interestingly, the MSIL2/4-associated proteins also include the phylogenetically related RBGD2/4 proteins, two constitutively expressed RBPs that are involved in heat resistance in Arabidopsis20. The comparison of the MSIL2/4 and RBGD2/4 interactomes indicates only a partial overlap, suggesting a certain level of subfunctionalization. In accordance with this observation, the MSILs and RBGDs clearly diversify in function, since the rbgd2/4 double mutant exhibits no developmental phenotypes under normal growth conditions.

A penetrant developmental phenotype associated with the MSIL2/4 mutations is a loss in stem rigidity that we could trace back to a defect in SCW formation, an important reservoir of fixed carbon and a renewable resource of major environmental and economic importance1. Yet, with the exception of few microRNAs involved in the control of lignin polymer synthesis6, 15, 16, no information about RBPs-mediated posttranscriptional regulation of SCW biosynthetic genes was previously reported. In this study, we show that MSIL2 and MSIL4, function redundantly to promote SCW formation in the interfascicular fibers and that their ability to bind RNA is essential to fulfill this function. Our conclusions are supported by independent approaches which highlight changes of SCW thickness, lignin content, and xylan decoration profile in the interfascicular fiber cells of the msil2/4 mutant. The fiber-specific effect of the MSIL2/4 mutations on SCW formation, is further supported by our evidence showing that the msil2/4 mutant grows normally and lacks the xylem collapses usually observed in more pleiotropic SCW deficient mutants. Previous studies have shown that specific master transcriptional regulators, such as NST1 and NST3, control cell-lineage specific formation of SCW formation in Arabidopsis5, 31. Our study extend this observation by showing that RBP-mediated post-transcriptional mechanisms are also involved in cell-type specific control of SCW formation in Arabidopsis. Notably, the presumed role of MSIL2/4 in interfascicular fiber cells contrasts with the rather constitutive expression of the MSIL2/4 genes, suggesting that their activity depends on cell-context. This is in accordance with the reported cell-type specific activities of ubiquitous animal MSI2 in cell renewal, cell differentiation and cancer control18, 19. Understanding how the activity of the MSIL2/4 genes is controlled by the cellular context will be essential from a basic point of view, providing a unique model for studying RBP regulation in specific cell type in plants.

Our data indicate that the accumulation of gene transcripts of SCW biosynthesis pathway enzymes and regulators are unaffected in msil2/4, suggesting that MSIL2/4 proteins are likely to act at a translational level to regulate the SCW formation in interfascicular fiber cells, an observation that is supported by the nature of the MSIL2/4 interactome. In this regard, the stem proteomics gives us a glimpse into the changes in protein abundance that could account for the observed msil2/4 SCW phenotype. Indeed, our MS data reveal that the glucuronoxylan biosynthesis machinery, including enzymes involved both in the synthesis and substitution of the xylan backbone, is specifically over-accumulated in the msil2/4 double mutant. The molecular specificity of the observed glucuronoxylan phenotype with respect to the changes in glucuronoxylan enzymatic machinery content raises several interesting questions about the impact of the MSIL2/4 mutations on the SCW formation. First, we noticed that the outcome of our proteomic data contrasts with the observed normal level of glucuronoxylan deposition in the interfascicular fibers SCW, suggesting that independent regulatory mechanisms act downstream of MSIL2/4 to control final glucuronoxylan content in the SCW. One possible explanation for this observation could be that the activities of the GUX, ESK, IRXs glucuronoxylan biosynthesis enzymes, in contrast to that of GXM3, are not rate-limiting for glucuronoxylan synthesis in vivo. Alternatively, and in agreement with studies that have reported that glucuronoxylan and cellulose contents are correlated in vivo in xylan-deficient mutants43, one could imagine that cellulose is the primary determinant of the content in glucuronoxylan in the SCW, and that the glucuronoxylan produced in excess in the golgi compartment of the msil2/4 mutant is targeted for degradation upon secretion if unbound to cellulose. In contrast, the concomitance between the glucuronoxylan hypermethylation and the GXM3 overaccumulation phenotypes observed in msil2/4 clearly support the previous assumption that the GXM-mediated methylation is a rate-limiting step in glucuronoxylan that can be overcome by GXM protein overexpression42. Our data support the idea that MSIL2/4 play a specific role in the developmental control of the GXM protein accumulation/activity, actively restraining the level of glucuronoxylan methylation in the Arabidopsis interfascicular fiber cells. We currently posit that MSIL2/4 could control the level of the GXM enzymes by directly regulating the translation of the GXM mRNAs or by indirectly regulating the translation of an yet unknown translational regulator of the glucuronoxylan biosynthesis pathway (XTRe) (Fig. 7). Future biochemical and functional studies will be necessary to precisely understand the mechanism by which MSIL2/4 proteins impact gene expression during SCW synthesis in Arabidopsis.

Model of MSIL-dependent control of glucuronoxylan methylation in Arabidopsis and its consequence for SCW architecture.

In the interfascicular fiber cells of Col-0, MSIL2/4 restrain the translation of the glucuronoxylan biosynthesis machinery, including the rate-limiting GXM3 enzyme. This activity would keep the level of glucuronoxylan methylation at an intermediate level, therefore providing a biochemical environment that favors the interactions between the glucuronoxylan and lignin polymers. In the interfascicular fiber cells of msil2/4 mutant, the translation of the glucuronoxylan biosynthesis machinery, including GXM3, is increased, leading to the deposition of an over- methylated form of glucuronoxylan that has detrimental effects on glucuronoxylan-lignin interaction and the SCW formation. In a non exclusive manner, MSIL2/4 could also have a positive role in lignin synthesis, whose defect in msil2/4 would lead to a decrease in lignin content. The model was inspired from Grantham et al37.

Our data indicate that, in addition to the reported change in glucuronoxylan methylation, the MSIL2/4 mutations is also associated with a specific decrease of lignin content in the SCW of the interfascicular fiber cells, whose origin remains unclear. On the basis of the previous observations, we envision two possible mechanisms by which the MSIL2/4 mutations could impact the lignin deposition in the fiber cells. In the first model, we propose that MSIL2/4 positively control the lignin synthesis pathway in the interfascicular cells by promoting the translation of lignin biosynthesis genes or that of a putative lignin translational activator (LTAc) (Fig. 7). In this regard, although no GO enrichment could be observed for lignin biosynthesis related genes in the proteomic analysis, two lignin biosynthetic enzymes, CAD4 and PAL2, are significantly down-accumulated in msil2/4 mutant (Supplementary Table 3), an observation that could account for the observed decrease in lignin content. However, previous published biochemical and genetic analyses do not support this conclusion, as the single cad4 and pal2 knock-out mutants display no significant alteration of lignin content due to genetic redundancy9, 44. Moreover, thio-acidolysis analyses showed that the ratio or degree of polymerization of the different lignin monomers remain unchanged in the msil2/4 mutant, reinforcing the idea that the lignin biosynthesis pathway is functional in this mutant, leaving open the question of the specificity of MSIL2/4 action on the lignin pathway (Fig.7). An alternative explanation for the decrease in lignin phenotype observed in the msil2/4 mutant would be that it results from an indirect cascade effect primarily triggered by the msil2/4-dependent changes in glucuronoxylan methylation. In this regard, it has been proposed that the 4-O-methylation of GlcA substituent results in the xylan being more hydrophobic, possibly impacting its non-covalent interaction with lignin and the capacity of lignin to assemble in the SCW41, 45, 46. Meanwhile, the hypermethylation of the carbon-4 hydroxyl of the GlcA side chains could affect the capacity of glucuronoxylan to form ether linkages to lignin (Fig. 7). Indeed, a study recently reported the existence of an ether linkage between a carbon-6 hydroxyl of a mannosyl residue of a galactoglucomannan hemicellulose and lignin in pine wood47. Although xylan has been proposed to bind covalently to lignin via ether bonds involving xylose or Arabinose residues, it remains unclear to date whether glucuronoxylan can form GlcA-dependent ether linkages with lignin, and the functional relevance of such linkages remains to be demonstrated48.

Methods

Plant material

Arabidopsis thaliana Col-0 and mutants msil1-1 (GABI_462A04), msil2-1 (Salk_066670), msil3-1 (Salk_002477) and msil4-1 (Salk_094167) obtained from the SALK Institute genomic analysis laboratory were used in this study. PAB2-RFP/PAB2R line has been provided by Cécile Bousquet-Antonelli (LGDP, Université Perpignan). The primers used to genotype lines are listed in Supplementary Table 4. Double mutants and quadruple mutants were generated by crossing plants. Plants were either grown in soil or cultivated in vitro on plates containing 2.20 g/l synthetic Murashige and Skoog (MS) (Duchefa) medium and 0.5g/l MES, pH 5.7, and 7 g/l agar. To breakdown dormancy, seeds were incubated for 48 h at 6°C in the dark. Germination and development were performed in growth chambers, at 20°C, 60–75% hygrometry with a 16-h light/8-h dark photoperiod (100 μE m −2 s−1 light [fluorescent bulbs with white 6500K spectrum, purchased from Sylvania]). For in vitro culture, the seeds were surface-sterilized before been sown on plates, incubated for 48 h at 6°C in the dark, and placed in a growth cabinet at 20°C with a 16-h-d/8-h-dark cycle and 120 μE m −2 s−1 light (LEDs with white 4500K spectrum, purchased from Vegeled).

Transgenic lines construction

MSIL2F-and MSIL4F-tagged versions were produced from Phusion-generated genomic DNA PCR products (promoter and coding region) using primers TL976(SalI)-TL924(PstI) for MSIL2F and TL974(SalI)-TL975(BamHI) for MSIL4F. PCR products were cloned in the CTL235 binary vector containing a Flag-HA tag in front of a NosT terminator and the hygromycin resistance gene driven by a 35S promoter. To obtain the MSIL2G and MSIL4G constructs used for fluorescence microscopy, the same PCR products were cloned in the CTL579 binary vector in front of a EGFP tag. The obtained plasmids were then used to transform mgl2/4 plants via Agrobacterium transformation. MSIL4- GFP/MSIL4G-tagged version used for complementation and RIP corresponds to the fusion of a genomic PCR product containing MSIL4 promoter (primers TL3527(HindIII)-TL3528(SalI)) fused with a second PCR cDNA fragment (primers TL3529(SalI)-TL3530(BamHI)) cloned initially in pGEM T Easy (Promega). After sequencing the fusion DNA has been cloned in the binary vector CTL579 containing GFP cDNA sequence. The MSIL4G-tagged construct was finally introduced in msil2/4 plants via Agrobacterium transformation. The rrm mutated MSIL4GΔr version was produced as MSIL4G except that the cDNA fragment has been gene-synthetized (Genecust) to modify RNA binding ability by having aspartic instead of phenylalanine at position 9, 49 and 51 in RRM1 domain and in position 111, 151 and 153 in RRM2 domain. Transgenic seeds selection was performed in the presence of 30 μg/l hygromycin and the progenies were screened by PCR.

Protein extraction and immunodetection

Total plant extracts (up to 100 mg) were ground in liquid nitrogen and proteins were extracted according to Hurkman and Tanaka49 Before migration, SDS–PAGE loading buffer was added and Coomassie staining was used to calibrate loadings. Proteins were separated on SDS/ PAGE gels and blotted onto Immobilon-P PVDF membrane (Cat. No. IPVH00010; MerckMillipore). Protein blot analysis was performed using the Immobilon Western Chemiluminescent HRP Substrate (Cat. No. WBKLS0500; MerckMillipore). Specific MSIL antibodies were raised in rabbits against defined epitopes (Epitope MSIL1 (SCDGTSSTFGYNRIPS), MSIL2 (RLQEYFGKYGDLVE), MSIL3 (GYGVKPEVRYSPAVGN) and MSIL4 (TWRSPTPETEGPAPFS) (Eurogentec).

Protein production and purification

The MSIL4 RRM binding domains were amplified using primers TL4128 and TL4129 either from MSIL4G construct for the wild type version or from MSIL4GΔr for the mutated version. PCR products were then cloned in pET41 to obtain His-tagged RRM domains. Tagged proteins were produced by induction in Escherichia coli BL21 cells. Cultures of 200 mL of an ampicillin-resistant (100 mg mL–1) colony were grown at 37°C and induced by 100mM isopropyl-b-D-galactopyranoside in the exponential phase (optical density 0.5 at 500 nm). After induction, bacteria were harvested by centrifugation and pellets were resuspended in 5 ml of binding buffer (His-bind buffer kit (Millipore), proteins were extracted using a Constant cell disrupter system (Constant Systems) with a disruption pressure of 2.35kbar and then purified using His-bind buffer kit (Millipore) following manufacturer’s recommendations.

RNA binding assays

Binding assays were performed using homopolymer RNA conjugated to CNBr-activated Sepharose beads (Sigma C9142). Either poly(A), poly(G), poly(C) or poly (U) from Mercks (respectively P9403, P4903, P4404 and P9528) were used following Sigma’s recommendations. 0.7 to 1 mg of RNA were used for 200 mg sepharose beads. 0.45 mg of purified protein (MSIL4 RRM binding domain or mutated version) in 0.5 ml binding buffer (100 mM NaCl, 10 mM Tris-HCl, pH 7.5, 2.5 mM MgCl2, 2.5 % Triton X-100) were incubated with 5 mg RNA conjugated Sepharose beads for 1 hour. After incubation, beads were washed 5 times for 10 min each in 0.5 ml of binding buffer. Beads were then pelleted at 12,000 rpm, resuspended in SDS sample buffer, boiled for 20 min, and pelleted again. Supernatants were transferred to fresh tubes and loaded on a 12% SDS-PAGE minigel for electrophoresis (Bio-Rad) and protein blot analysis.

RNA immunoprecipitation

RIP was performed as described previously in Merret et al.50 using 400 mg of stem powder as starting material.

Expression analysis

Total RNA was isolated using the TRI re-agent (Cat. No. TR-118; Euromedex) according to the manufacturer’s instructions. Genomic DNA was then digested out using the RQ1 RNase-Free DNase (Cat. No. M6101; Promega). cDNAs were obtained from 400 ng of RQ1-treated RNA using 1 μl of GoScript Reverse Transcriptase (Cat. No. A5003; Promega) in a 20-μl final volume reaction using random primers (Cat. No. C1181; Promega) in the presence of 20 units of RNasin (Recombinant Ribonuclease Inhibitor, Cat. No. C2511; Promega) and dNTP mix at a final concentration of 0.5 mM of each dNTP (Cat. No. U1511; Promega). Semi-quantitative RT–PCR amplifications were performed on 1 μl of cDNA in a 12.5-μl reaction volume to start with. The amplification of EF1αtranscripts was used to equilibrate. qRT–PCR was performed on a Light Cycler 480 II machine (Roche Diagnostics) by using the Takyon No ROX SYBR MasterMix blue dTTP kit (Cat. No. UF-NSMT-B0701; Eurogentec). Each amplification reaction was set up in a 10-μl reaction containing each primer at 300 nM and 1 μl of RT template in the case of total RNA with a thermal profile of 95°C for 10 min and 40 amplification cycles of 95°C, 15 s; 60°C, 60 s. Relative transcript accumulation was calculated using the ΔΔCt methodology51, using ACTIN2 as internal control. Average ΔΔCt represents three experimental replicates with standard errors. PolyA+ RNAs were purified from RNA obtained by the TRI Reagent extraction, using the PolyATtract mRNA Isolation System III (Cat. No. Z5310; Promega) according to the manufacturer’s instructions. cDNAs were then synthetized according to the above protocol using 50 ng of polyA+ RNA as starting material. Semiquantitative and quantitative RT–PCRs were performed as described above with a 1/25 dilution of cDNAs from polyA+ RNA as template for qRT–PCR. The primers used for RT and qRT-PCR are listed in Supplementary Table 6.

RNA Sequencing

Total RNA were extracted from the basal stem section of mature Col-0 and msil2/4 mutant plants using the RNeasy Plant Mini Kit (Qiagen, Cat.#74904) and treated onto the column with the RNase-Free DNase Set (Qiagen, Cat.#79254). Two replicates were performed per plant lines. PolyA purification, library preparation using the Illumina TruSeq stranded mRNA kit and library quality controls were performed by the BioEnvironment Illumina Platform. The mRNA sequencing of Col-0 and msil2/4 mRNAs was done with single-reads, 1x 75pb and 30–35×106 reads per library.

RNASeq analysis

For each library, 30–35×106 reads were obtained with 85–90% of the bases displaying a Q-score ≥ 30, with a mean Q-score of 38 as assessed with Fastqc (http://www.bioinformatics.babraham.ac.uk/projects/fastqc/). Reads were mapped against the TAIR 10 genome using gtf annotation file and default parameters of TopHat2 v2.0.752. Assembly and transcript quantification were performed with Cufflinks v2.2.153. Finally, the low expressed transcripts -less than 1 RPKM (reads per kilobase per million mapped reads) in one of the libraries were filtered out. The differential analysis was conducted by Cuffdiff software, belonging to the Trapnell suite. Foldchange (FC) was determined as the ratio between the normalized read counts between mutant and wild-type.

Fluorescence detection and tissue colorations

GFP detections were performed on in vitro grown one week seedling roots fixed in fixation buffer (50mM Pipes, 5mM EGTA, 5mM MgSO4, pH7 added with 2% para-formaldehyde and 0.2% Triton). After 5 min vacuum infiltration and 10 min agitation at room temperature, roots were washed twice with fixation buffer and stored at 4°C until confocal observation. Confocal imaging was done using an Zeiss Axio Observer Z1 LSM 700. Roots were observed in water and for GFP detection samples were excited at 488 nm and the signal captured at 495–555 nm. For RFP detection, excitation was performed at 555 nm and the signal captures at 600-700 nm. Images were analysed with ImageJ. At least 15 samples per line were analyzed.

Stem coloration were performed from 5 cm length basal stem pieces conserved in ethanol 80%. Stems were rehydrated first in ethanol 50% and later in water and then embedded in agarose 5% before being transversally cut using a vibratome (Leica vt 1000s) to obtain 120 micron thickness sections stored in water. The presence of lignin was determined by staining with phloroglucinol for 30 seconds, giving a red coloration in the presence of lignin cinnamaldehyde groups. Sections were observed using an inverted microscope (Leitz DMRIBE, Leica Micro-systems, Wetzlar, Germany) and images were registered using a CCD camera (Color Collview, Photonic Science, Milham, UK).

For cellulose detection, the brightener Calcofluor White ST [4,4’-bis(anilino-6-bis(2-hydroxyethyl)-amino-s-triazine-ylamino)-2,2’-stilbene disulfonic acid; Cyanamide Co., Bound Brook, N.J] was used at a final concentration of 0.1%. The samples were incubated with calcofluor for 3 min, washed three times with water before being observed. For detection, excitation was performed at 405 nm and the signal captures at 434-496 nm. For auto-fluorescence analysis stem sections were excited at 405 nm and the signal captures at 410-470 nm for blue. Pictures have been analyzed with LAS X (Leica Application Suite X).

Biochemical analyses of SCWs

Arabidopsis stems (5 cm sections at the bottom part of the stems) were harvested from Col0 and msil2/4 double mutant as four independent biological replicates. Soluble extractives were eliminated as described by Ployet et al.54. Briefly, stems were freeze dried for 48h, ground using a Mixer Mill MM 400 (Retsch), and extracted by hot solvents (successively water, ethanol, ethanol/toluene (1:1 v/v) and acetone) to obtain extractive-free stem residues (ESR). The determination of acetyl bromide (AcBr) lignin was performed on 10mg of ESR as described by Ployet et al.55. All analyses were done in triplicate. Fourier transform infrared spectroscopy (FT-IR) analysis was performed on 100-200 mg of ESR. Spectra were recorded from ten technical replicates, in the range 400–4000 cm−1 with a 4 cm-1 resolution and 32 scans per spectrum using an attenuated total reflection (ATR) Nicolet 6700 FT-IR spectrometer (Thermo Fisher, Illkirch-Graffenstaden, France) equipped with a deuterated-triglycine sulfate (DTGS) detector. Spectra analyses were performed as described in Dai et al.56. Saccharification yield was estimated with or without alkali pretreatment, as described by Van Acker et al.57. Reducing sugar concentration was assessed with dinitrosalicylic acid (DNS) reagent using 10µl of the supernatant after 9 h of reaction. Enzyme activity was assessed at 0.25 filter paper units (FPU) ml-1 in our conditions.

Mass spectrometry (MS)-based proteomic analyses

Proteins from total inflorescence stem extracts of three biological replicates of Col-0 and msil2/4 mutant plants were solubilized in Laemmli buffer and heated for 10 min at 95°C. They were then stacked in the top of a 4-12% NuPAGE gel (Invitrogen), stained with Coomassie blue R-250 (Bio-Rad) before in-gel digestion using modified trypsin (Promega, sequencing grade) as previously described58. The resulting peptides were analyzed by online nanoliquid chromatography coupled to MS/MS (Ultimate 3000 RSLCnano and Orbitrap Exploris 480, Thermo Fisher Scientific) using a 180- min gradient. For this purpose, the peptides were sampled on a precolumn (300 μm x 5 mm PepMap C18, Thermo Scientific) and separated in a 75 μm x 250 mm C18 column (Aurora Generation 2, 1.6µm, IonOpticks). The MS and MS/MS data were acquired using Xcalibur 4.4 (Thermo Fisher Scientific).

Peptides and proteins were identified by Mascot (version 2.8.0, Matrix Science) through concomitant searches against the Viridiplantae database (from Uniprot, Arabidopsis thaliana (thale-cress) taxonomy, 136447 sequences) and a homemade database containing the sequences of classical contaminant proteins found in proteomic analyses (human keratins, trypsin…, 126 sequences). Trypsin/P was chosen as the enzyme and two missed cleavages were allowed. Precursor and fragment mass error tolerances were set at respectively at 10 and 20 ppm. Peptide modifications allowed during the search were: Carbamidomethyl (C, fixed), Acetyl (Protein N-term, variable) and Oxidation (M, variable). The Proline software59 (version 2.2.0) was used for the compilation, grouping, and filtering of the results (conservation of rank 1 peptides, peptide length ≥ 6 amino acids, false discovery rate of peptide-spectrum-match identifications < 1%60, and minimum of one specific peptide per identified protein group). MS data have been deposited to the ProteomeXchange Consortium via the PRIDE partner repository61. Proline was then used to perform a MS1 label-free quantification of the identified protein groups based on razor and specific peptides.

Statistical analysis was performed using the ProStaR software62 based on the quantitative data obtained with the four biological replicates analyzed per condition. Proteins identified in the contaminant database, proteins identified by MS/MS in less than two replicates of one condition, and proteins quantified in less than three replicates of one condition were discarded. After log2 transformation, abundance values were normalized using the variance stabilizing normalization (vsn) method, before missing value imputation (SLSA algorithm for partially observed values in the condition and DetQuantile algorithm for totally absent values in the condition). Statistical testing was conducted with limma, whereby differentially expressed proteins were selected using a log2(Fold Change) cut-off of 1 and a p- value cut-off of 0.00513, allowing to reach a false discovery rate close to 1% according to the Benjamini-Hochberg estimator. Proteins found differentially abundant but identified by MS/MS in less than two replicates, and detected in less than three replicates, in the condition in which they were found to be more abundant were invalidated (p-value = 1).

MALDI-TOF MS analyses of the patterns of xylanase-released SCW oligosaccharide

10 mg of dried alcohol insoluble cell wall residues were incubated with 10U.ml-1 of endo-1,4-D- xylanase (E-XYLNP) from MEGAZYME previously desalted with centrifulgal filters (Amicon Ultra-0.5ml) in 1 ml during 12hrs at 40°C. After boiling during 5 min to inactivate enzyme activity, samples where filtred with 50KD centrifulgal filters (Amicon Ultra-0.5ml). As positive control, 2 mg of wheat AX (MEGAZYMZE) were treated in a similar manner. The samples were then analyzed by matrix-assisted laser desorption ionization-time of flight (MALDI-TOF) MS using DMA/DHB (N,N-dimethylaniline/2,5-dihydroxybenzoic acid) matrix39. The samples (1 μL) were deposited and then covered by the matrix (1 μL) on a polished steel MALDI target plate. MALDI measurements were then performed on a rapifleX MALDI-TOF spectrometer (Bruker Daltonics, Bremen, Germany) equipped with a Smartbeam 3D laser (355 nm, 10000 Hz) and controlled using the Flex Control 4.0 software package. The mass spectrometer was operated with positive polarity in reflectron mode. Spectra were acquired in the range of 560–3200 m/z. The detected species correspond to ions in the form of sodium adducts.

Thioacidolysis

This method is adapted from Lapierre et al.63. Briefly, 10 mg of alcohol insoluble cell wall residues were incubated in 3ml of dioxane with ethanethiol (10%), BF3 etherate (2.5%) containing 0.1% of heinecosane C21 diluted in CH2Cl2 at 100°C during 4 hours. 3ml of NaHCO3 (0.2 M) were added after cooling and mixed prior the addition of 0.1 mL of HCl (6 M). The tubes were wortexed after addition of 3 mL of dichloromethane and the whole lower organic phase was collected in a new tube before concentration under nitrogen atmosphere to approximately 0.5 ml. Then, 10 μL were trimethylsilylated (TMS) with 100 μL of N,O-bis(trimethylsilyl) trifluoroacetamide and 10 μL of ACS-grade pyridine. The trimethylsilylated samples were injected (1 μL) onto an Agilent 5973 Gas Chromatography–Mass Spectrometry system. Specific ion chromatograms reconstructed at m/z 239, 269 and 299 were used to quantify H, G and S lignin monomers respectively and compared to the internal standard at m/z 57, 71, 85).

Cell wall sugar analysis

Neutral monosaccharide sugar content was determined by gas chromatography after acid hydrolysis and conversion of monomers into alditol acetates as described in Hoebler et al.64 and Blakeney et al.65 Gas Chromatography was performed on a DB 225 capillary column (J&W Scientific, Folsorn, CA, USA; temperature 205 °C, carrier gas H2). Calibration was made with standard sugar solution and inositol as internal standard.

For acid sugar analysis, the uronic acid content was automatically quantified after hydrolysis of polysaccharides in concentrated sulfuric acid (18 M) containing sodium tetraborate (12.5 mM) or not, followed by m-hydroxydiphenyl colorimetric determination66, 67. Galacturonic and Glucuronic acid were used as standard for quantification.

Data availability

The high-throughput sequencing data generated in this study are accessible at NCBI’s Gene Expression Omnibus (GEO) via GEO Series accession number GSE223732. The mass spectrometry proteomics data have been deposited to the ProteomeXchange Consortium via the PRIDE partner repository with the dataset identifier PXD040207 (proteomic identification of MSIL-interacting partners), and PXD040020 (proteomic analysis of total inflorescence stem extracts).

Supporting information

eLife Assessment

Secondary cell walls support vascular plants and conduct water throughout the plant body, and are important resources for lignocellulosic feedstocks. Here the authors present convincing genetic and biochemical evidence that secondary cell wall synthesis, known already to be under complex transcriptional control, is also controlled post-transcriptionally by MUSASHI-like RNA-binding proteins. These important results point to a new mechanism for control of secondary cell wall synthesis, which will be interesting to cell biologists and biochemists studying and attempting to manipulate plant biomass.

Reviewer #1 (Public Review):

Secondary cell walls support vascular plants and conduct water throughout the plant body, but are also important resources for lignocellulosic feedstocks. Secondary cell wall synthesis is under complex transcriptional control, presumably because it must only be initiated after cell growth is complete. Here, the authors found that two Musashi-type RNA-binding proteins, MSIL2 and MSIL4 are redundantly required for secondary cell wall development in Arabidopsis. The plant phenotypes could be complemented by the wild-type version of either protein, but not by a MSIL4 version that carries mutations in the conserved RNA-binding domains, and the authors localized MSIL2 & 4 to stress granules, implicating the RNA-binding function of MSIL4 in the cell wall phenotype. Upon closer inspection, the secondary cell wall phenotypes included changes in vasculature morphology, and minor changes to lignin and hemicellulose (glucuronoxylan). While there were no changes to likely cell wall target genes in the transcriptome of msil2msil4 plants, proteomics experiments found glucuronoxylan biosynthesis components were upregulated in the mutants, and they detected an increase in substituted xylan via several methods. Finally, they documented MSIL4 binding to RNA encoding one of these targets, suggesting that MSIL2 and MSIL4 act to post-transcriptionally regulate glucuronoxylan modification. Altogether, this is a new mechanism by which cell wall composition could be regulated.

Overall, the manuscript is well-written, the data are generally high-quality, and the authors typically use several independent methods to support each claim. However, several important questions remain unanswered by this work in its current state and the model presented in Figure 7 is quite speculative. For example, the link between the striking plant phenotype and GXM misregulation is unclear since GXM overexpression doesn't alter plant phenotypes or lignin content (Yuan et al 2014 Plant Science), so misregulation of GXMs in msil2msil4 mutants clearly is not the whole story. It also remains to be determined why one particular secondary cell wall synthesis enzyme is regulated likely post-transcriptionally, while so much of the pathway is regulated at the transcriptional level. There are likely other targets for MSIL2- and MSIL4-mediated regulation since it seems that MSIL2 and MSIL4 are expressed in tissues that are not synthesizing secondary cell walls.

Reviewer #2 (Public Review):

This work explored the biological functions of a small family of RNA-binding proteins that was previously studied in animals, but was uncharacterized in plants. Combinatorial T-DNA insertional mutants disrupting the expression of the four Mushashi-like (MSIL) genes in Arabidopsis revealed that only the msil2 msil4 double mutant visibly alters plant development. The msil2/4 plants produced stems that could not stand upright. Transgene complementation, site-directed mutagenesis of MSIL4 conserved RNA-binding motifs, and in vitro RNA binding assays support the conclusion that the loss of MSIL2 and MISL4 function is responsible for the observed morphological defects. MSIL2/4 interact with proteins associated with mRNA 3'UTR binding and translational regulation.

The authors present compelling biochemical evidence that Mushashi-like2 (MSIL2) and MSIL4 jointly regulate secondary cell wall biosynthesis in the Arabidopsis stem. Quantitative analyses of proteins and transcripts in msil2/4 stems uncovered upregulation of several xylan-related enzymes (despite WT-like RNA levels). Consistent with MALDI-TOF data for released xylan oligosaccharides, the authors propose a model in which MSIL2/4 negatively regulate the translation of GXM (glucuronoxylan methyltransferase), a presumed rate-limiting step. The molecular links between overmethylated xylans and the observed stem defects (which include subtle reductions in lignin and increases beta-glucan polymer distribution) warrants further investigation in future studies. Similarly, as the authors point out, it is intriguing that the loss of the broadly expressed MSIL2/4 genes only significantly affects specific cell types in the stem.

Reviewer #3 (Public Review):

The manuscript by Kairouani et al. investigates the function of a small family of plant RNA binding proteins with similarity to the well-studied Musashi protein in animals, and, therefore, called MUSASHI-LIKE1-4 (MSL1-4). Studies on the biological importance of post-transcriptional control of gene expression via RNA-binding proteins in plants are not numerous, and advances in this important field are much needed. The thorough work presented in this manuscript is such an advance.

The central observations of the paper are:

- Knockout of any MSL gene alone does not produce a phenotype.

It is of note that basic characterization of knockout mutations is really well done - for example, the authors have taken care to raise specific antibodies to each of the MSL proteins and use them to demonstrate that each of the T-DNA insertion mutants used actually does knock out protein production from the corresponding gene.

- Knockout of MSL2/4 (but no other double mutant) produces a clear leaf phenotype, and a remarkable stem phenotype in which the mutants collapse as they are unable to support upright growth

- The phenotypes of knockout mutants persist in point mutants defective in RNA-binding, indicating that RNA-binding is required for biological activity. Consistent with this, and associate physically with other RNA-binding proteins and translation factors.

- MSL proteins are cytoplasmic

- The msl2/4 mutants present multiple defects in secondary cell wall composition and structure, probably explaining their inability to grow upright. I did not examine the cell wall analyses in detail as I am no specialist in this field.

- Msl2/4 mutants show transcriptomic changes with at large two big categories of differentially expressed genes compared to wild type.

(1) Genes related to cell wall metabolism

(2) Genes associated with defense against herbivores and pathogens

- Two of the mRNAs encoding cell wall factors with significant upregulation in msl2/4 mutants compared to wild type also associate physically with MSL4 as judged by RNA-immunoprecipitation-RT-PCR assays, and this physical association is abrogated in the RNA-binding deficient MSL4 mutant.

Altogether, the study shows clear biological relevance of the MSL family of RNA-binding proteins, and provides good arguments that the underlying mechanism is control of mRNAs encoding enzymes involved in secondary cell wall metabolism (although concluding on translational control in the abstract is perhaps saying too much - post-transcriptional control will do given the evidence presented). One observation reported in the study makes it vulnerable to alternative interpretation, however, and I think this should be explicitly treated in the discussion:

The fact that immune responses are switched on in msl2/4 mutants could also mean that MSL2/4 have biological functions unrelated to cell wall metabolism in wild type plants, and that cell wall defects arise solely as an indirect effect of immune activation (that is known to involve changes in expression of many cell wall-modifying enzymes and components such as pectin methylesterases, xyloglucan endotransglycosylases, arabinogalactan proteins etc. Indeed, the literature is rich in examples of gene functions that have been misinterpreted on the basis of knockout studies because constitutive defense activation mediated by immune receptors was not taken into account (see for example Lolle et al., 2017, Cell Host & Microbe 21, 518-529).

With the evidence presented here, I am actually close to being convinced that the primary defect of msl2/msl4 mutants is directly related to altered cell wall metabolism, and that defense responses arise as a consequence of that, not the other way round. But I do not think that the reverse scenario can be formally excluded with the evidence at hand, and a discussion listing arguments in favor of the direct effect proposed here would be appropriate. Elements that the authors could consider to include would be the isolation of a cellulose synthase mutant as a constitutive expressor of jasmonic acid responses (cev1) as a clear example that a primary defect in cell wall metabolism can produce defense activation as secondary effect. The interaction of MSL4 with GXM1/3 mRNAs is also helpful to argue for a direct effect, and it would strengthen the argument if more examples of this kind could be included.

Author Response

The following is the authors’ response to the original reviews.

Please find below our detailed point-by-point response to the eLife reviewer comments. As suggested by the reviewers, we have 1) replaced most of the Bar charts by Box plots, 2) highlighted the sucellular regions that are analyzed in the measurement experiments, and 3) have rewritten and toned down several subsections of the discussion.

Reviewer #1 (Recommendations For The Authors):

I suggest that the authors consider the following points in future versions of this manuscript:

1). The link between the striking plant phenotype and GXM misregulation is unclear since GXM overexpression doesn't alter plant phenotypes or lignin content (Yuan et al 2014 Plant Science), so misregulation of GXMs in msil2msil4 mutants clearly is not the whole story. The authors should discuss alternative interpretations of their results and other possible targets of MSIL2/4 that might be contributing to the plant phenotype.

We completely agree with the reviewer that the misregulation of GXMs in msil2/4 is not the whole story and we are currently developing specific strategies in order to characterize in an unbiased manner the full repertoire of MSIL mRNA targets in the stem, hoping we can identify other targets relevant to the formation of SCW. We have also toned-down our discussion concerning the possible impact of glucuronoxylan methylation level on lignin deposition (L546-552).

1. Similarly, it remains unclear why one particular secondary cell wall enzyme is regulated post-transcriptionally, while so much of the pathway is regulated at the transcriptional level. Please discuss.

We do not exclude that other genes encoding for SCW enzymes are impacted and it will be the subject of further investigations. We have extended the discussion concerning these points. We have extended the discussion concerning these points (L486-498).

1. Thirdly, it seems that MSIL2 and MSIL4 are expressed in tissues that are not synthesizing secondary cell walls. The authors should discuss other possible targets of MSIL2/4 from their work.

We have extended the discussion concerning the pleiotropic effects of MSIL mutation in Arabidopsis (L 416-425). The variability of the msil2/4 phenotype is so large that we expect these proteins to regulate various cellular functions through the binding of specific set of mRNA. The mRNA targets specifically involved in these regulations will need to be determined on a case-by-case basis.

1. The discussion is extremely speculative and introduces new abbreviations (LTAc, XTRe) that are only used in their model (Figure 7). I suggest replacing these with dashed lines and/or question marks in the model, since as currently depicted, it looks as if these could be known gene products, which could be very misleading.

We have removed the Ltac and XTRe abbreviations in Figure 7, and the corresponding text in the discussion section.

1. Similarly, the speculation that cellulose content somehow regulates glucuronoxylan levels via xylan-cellulose interactions, leading to degradation of excess glucuronoxylan after synthesis is, to my knowledge, completely unsupported by any evidence except the correlation between cellulose and xylan levels. Please either support this claim with references or remove it from the discussion.

We have removed the claim and have rewritten and toned down the text accordingly to the reviewer 1 comments (L 499-512).

1. Bar charts are rarely the most appropriate method for displaying biological data (Streit & Gehlenborg 2014 Nature Methods). Authors should replace bar charts with one of the following options: A) plot all individual datapoints and overlay summary statistics, B) box plots with all individual datapoints show, C) violin plots (when n is large, i.e. n > 50). R and R studio are free software that can generate such plots. Several excellent tools exist online to generate such plots via a free, graphical user interface, such as boxplotr (Spitzer et al 2014 Nature Methods): http://shiny.chemgrid.org/boxplotr/ and PlotsOfData (Postma & Goedhart 2019 PLoS Biology): https://huygens.science.uva.nl/PlotsOfData/

We have replaced the Bar charts in figure 4E,G and Fig 5E with Box plots and acknowledged the software used in the corresponding Materials and methods section.

Reviewer #2 (Recommendations For The Authors):

Minor points:

Which cells from Fig. 4b were measured for 4c? Some highlighted annotations to delineate the regions that were measured would help.

We have highlighted in figure 4B the subcellular regions cells analyzed in the measurement experiments.

In line 254, the phrase "not merely affected" in the mutant should be rephrased for clarity

We have replaced “not merely affected” by “not significantly” (L274).

Line 317: "we first performed glycome profiling", the data shows monosaccharide profile, not glycome profiling usually involving antibodies microarrays

We have corrected the text according to the reviewer comment (L339-340).

Reviewer #3 (Recommendations For The Authors):

Altogether, the study shows clear biological relevance of the MSL family of RNA-binding proteins, and provides good arguments that the underlying mechanism is control of mRNAs encoding enzymes involved in secondary cell wall metabolism (although concluding on translational control in the abstract is perhaps saying too much - post-transcriptional control will do given the evidence presented). One observation reported in the study makes it vulnerable to alternative interpretation, however, and I think this should be explicitly treated in the discussion:

The fact that immune responses are switched on in msl2/4 mutants could also mean that MSL2/4 have biological functions unrelated to cell wall metabolism in wild type plants, and that cell wall defects arise solely as an indirect effect of immune activation (that is known to involve changes in expression of many cell wall-modifying enzymes and components such as pectin methylesterases, xyloglucan endotransglycosylases, arabinogalactan proteins etc. Indeed, the literature is rich in examples of gene functions that have been misinterpreted on the basis of knockout studies because constitutive defense activation mediated by immune receptors was not taken into account (see for example Lolle et al., 2017, Cell Host & Microbe 21, 518-529).

With the evidence presented here, I am actually close to being convinced that the primary defect of msl2/msl4 mutants is directly related to altered cell wall metabolism, and that defense responses arise as a consequence of that, not the other way round. But I do not think that the reverse scenario can be formally excluded with the evidence at hand, and a discussion listing arguments in favor of the direct effect proposed here would be appropriate. Elements that the authors could consider to include would be the isolation of a cellulose synthase mutant as a constitutive expressor of jasmonic acid responses (cev1) as a clear example that a primary defect in cell wall metabolism can produce defense activation as secondary effect. The interaction of MSL4 with GXM1/3 mRNAs is also helpful to argue for a direct effect, and it would strengthen the argument if more examples of this kind could be included.

In accordance to Rev3 comments, we have extended the discussion, listing the arguments, that we believe, are not in favor of a primary effect of the MSIL2/4 proteins on the activation of plant defense pathways (L468-485).

SUGGESTIONS FOR IMPROVED ANALYSES & MINOR TEXT AND FIGURE CORRECTIONS.

(1) Unless there is a very good reason to use homology modelling such as SWISS-MODEL (for example ligand-bound proteins), Alphafold2 is now the tool to use for structure prediction. I would at least verify that Alphafold agrees with SWISS-MODEL on the predicted structures shown in Fig 2a.

We have analyzed the MSIL4 sequence using the Alphafold2 prediction software and the output of this analysis completely agrees with the SWISS-Model prediction. We have added an additional panel showing the Alphafold 2 prediction (see figure 2-figure supplement 1B).

(2) The plant pictures shown in Figure 2d are not publication quality in terms of resolution, mounting, size. They really should be redone before final publication.

We thank the reviewer for this important observation, and have improved the resolution of the figure 2D.

(3) The colocalization in Figure 3d/e would benefit from some statistical analysis of the data: How many foci were examined? How many showed colocalization? Is that fraction statistically significant? It can be done from the images at hand; I do not think that additional data acquisition is necessary.

We have used an ImageJ plugin to perform colocalization analysis on the microscopy images corresponding to the bottom panel of the figure 3D (heat stress). This analysis confirmed that most of the foci are actually colocalizing (see Author response image 1). However our initial image data acquisition do not allow us to perform statistical analysis on it. We have added a sentence indicating that colocalization is supported by an analysis using an ImageJ plugin.

1. Typographical and other writing errors:

Line 72 "prior to"

Line 77 "in the Arabidopsis model"

Line 97 "RBP-mediated..."

Line 110 "aspects of development"

Line 128 "little is known" (no yet)

Line 253 "Col-0"

Line 346 "previous"

All the writing errors have been corrected in the revised version.

Version 2

Abstract

Deciphering the mechanism of secondary cell wall/SCW formation in plants is key to understanding their development and the molecular basis of biomass recalcitrance. Although transcriptional regulation is essential for SCW formation, little is known about the implication of post-transcriptional mechanisms in this process. Here we report that two bonafide RNA-binding proteins homologous to the animal translational regulator Musashi, MSIL2 and MSIL4, function redundantly to control SCW formation in Arabidopsis. MSIL2/4 interactomes are similar and enriched in proteins involved in mRNA binding and translational regulation. MSIL2/4 mutations alter SCW formation in the fibers, leading to a reduction in lignin deposition, and an increase of 4-O-glucuronoxylan methylation. In accordance, quantitative proteomics of stems reveal an overaccumulation of glucuronoxylan biosynthetic machinery, including GXM3, in the msil2/4 mutant stem. We showed that MSIL4 immunoprecipitates GXM mRNAs, suggesting a novel aspect of SCW regulation, linking post-transcriptional control to the regulation of SCW biosynthesis genes.

Introduction

As the main constituent of the terrestrial plant biomass, lignocellulose represents a huge reservoir of fixed carbon and a renewable resource for bioproducts and energy that is essential for human usages (Meents et al., 2018). Lignocellulose biomass enclosed mainly within secondary cell wall (SCW) reinforces vessels for long-distance transport and provides physical properties to fibers allowing upright growth of the plants (Meents et al., 2018; Scheller and Ulvskov, 2010). SCWs are composite material made of high molecular weight biopolymers, including cellulose, hemicelluloses, and lignin, as well as cell wall proteins whose composition and structure can differ markedly between dicots and monocots (Scheller and Ulvskov, 2010; Yokoyama and Nishitani, 2004). Cellulose is synthesized at the plasma membrane by the cellulose synthase complex and is composed of linear chains of β-(1–4) glucans that aggregate to form highly crystalline cellulose microfibrils (Scheller and Ulvskov, 2010; Yokoyama and Nishitani, 2004). Xylans, the main hemicellulosic polysaccharides in SCW, are synthesized in the Golgi apparatus by a complex biosynthetic machinery and comprise a group of polysaccharides that share a common backbone of β-(1,4)-linked xylose (Xyl) units, but differs by the presence of side chains, whose nature is dependent on tissues and species (Scheller and Ulvskov, 2010). In particular, glucuronoxylan, the dominant xylan in the SCW of dicots, is notably substituted at O-2 position with α-D-glucuronic acid (GlcA) or (4-O-methyl)-α-D-glucuronic acid (MeGlcA) groups, and with O-acetyl moieties at C-2 or C-3 positions (Scheller and Ulvskov, 2010). By interacting with the cellulose and lignin, xylans contribute to the strengthening of the SCW, and are considered to be one of the main factors contributing to the biomass recalcitrance to enzymatic hydrolysis (Yokoyama and Nishitani, 2004; Kumar et al., 2016). Finally, lignin polymers are made up of H, G, and S units which are the results of the oxidation and the polymerisation of the p-coumaryl, coniferyl and Synapyl alcohols, respectively. These monomers, also called monolignols are synthesized on the ER surface/cytoplasm prior to transportation to the cell wall. Lignins confer stiffness, strength and hydrophobicity to the SCW (Scheller and Ulvskov, 2010; Yokoyama and Nishitani, 2004; Kumar et al., 2016).

The metabolic load imposed by the coordinated production of SCW biopolymers emphasizes the importance for plants of having a precise control over temporal and spatial expression of the corresponding SCW biosynthetic genes. Based on founding studies performed in the Arabidopsis model, it is generally recognized that the SCW synthesis is primarily controlled at a transcriptional level by a multilayer and interconnected network of transcription factors/TFs (Taylor-Teeples et al., 2015; Zhang et al., 2018). Key players of this regulatory network are related NAC (NAM-ATAF1/2-CUC) and MYB-type TFs that induce, in a redundant and combinatorial manner, the expression of SCW biosynthetic genes by binding their promoter region (Taylor-Teeples et al., 2015). In addition, a family of Class III homeodomain leucine zipper (HD-ZIPIII) TFs, whose members control several aspects of Arabidopsis development, was also shown to interact with SCW regulatory network, contributing to xylem cell specification and SCW synthesis (Taylor-Teeples et al., 2015; Zhang et al., 2018). Although this transcriptional level of control can gain complexity in perennial plants, functional orthologs of the main regulators were identified in many different species, suggesting that SCW deposition proceeds via a conserved mechanism in all vascular plants (Myburg et al., 2014; Zhong and Ye, 2014). During the last decade, engineering the SCW regulatory network to deposit modified lignin and/or improve polysaccharide composition became a promising strategy to optimize biomass processability (Sibout et al., 2005; Whitehead et al., 2018). However, despite huge progresses in understanding the regulation of SCW synthesis, the engineering of SCW regulatory network to improve biomass yield and digestibility has proven difficult and often impairs plant growth because of altered vascular tissue development (Bonawitz and Chapple, 2013; Wang et al., 2016).

Downstream of transcriptional regulations by TFs, RNA-binding proteins (RBPs) are essential post-transcriptional modulators of gene expression across all kingdoms of life (Singh et al., 2015). However, although a large cohort of RBPs exists in plants (Reichel et al., 2016), there is little information about RBP-mediated posttranscriptional regulation of SCW biosynthetic genes. So far, only few microRNAs (miRNAs) families have been implicated in the post-transcriptional regulation of genes involved in both regulatory and enzymatic aspects of SCW biosynthesis (Zhang et al., 2018). In particular, miR165/166 were shown to target HD-ZIPIII transcriptional regulators, having a direct impact on the transcriptional network associated with SCW biosynthesis (Zhang et al., 2018). In the other hand, miRNA397/857/408 were shown to affect lignin content via the targeting of laccase/LAC genes, which control lignin polymerization from monolignols precursors (Wang et al., 2014; Zhao et al., 2015). In addition, two related tandem CCCH zinc-finger proteins, which exhibit both DNA and RNA binding abilities, have been proposed to modulate SCW formation by regulating the expression of genes associated with cell wall metabolism (Chai et al., 2015). Despite these mounting evidences, our current knowledge of the role of post-transcriptional regulators in SCW synthesis is still very limited.

In this work, we report that two RRM-domain containing RNA-binding proteins homologous to the animal translational regulator Musashi, Musashi-like2/MSIL2 and Musashi-like4/MSIL4, function redundantly to control various aspects of development, including the stiffness of the inflorescence stem in Arabidopsis. We show that RRM-dependent RNA-binding activity is essential for MSIL2/4 functions in vivo, and that MSIL2/4 interactomes are similar, being enriched in proteins involved in 3’-UTR binding and translational regulation. MSIL2/4 mutations alter the formation of SCW in the interfascicular fiber cells, leading to a reduction in lignin deposition, and a change in the decoration pattern of glucuronoxylan that is associated with an increase of 4-O-methylation of GlcA substituent. In accordance, quantitative mass-spectrometry-based protein analysis reveals an overaccumulation of glucuronoxylan biosynthetic machinery, including the GlucuronoXylan Methyltransferase3/GXM3, in the msil2/4 mutant stem. We show that MSIL4 immunoprecipitates GXM3 mRNA in vivo, likely regulating its expression at a translational level. Our results demonstrate that MSILs regulate SCW synthesis in interfascicular fiber cells and point to a novel aspect of SCW regulation linking translational repression to regulation of SCW biosynthesis genes.

Results

Musashi-like MSIL2/4 proteins redundantly control specific Arabidopsis development processes

While many studies have implicated the MSIs in the control of gene expression during cellular proliferation, cell fate determination and cancer in animals (Fox et al., 2015; Kudinov et al., 2017), little is known about the presence and activity of MSI-type proteins in plants. In this work, we describe an RNA-binding protein family in Arabidopsis, whose members hereafter named as MUSASHI-Like1 to 4/MSIL1-4, share both sequence, domain organization, and model-based structural similarities with animal MSIs (Figure 1A and B). Phylogenetic analysis revealed that Arabidopsis indeed harbors a clade of seven MSIL-type genes with a prominent sub-clade harboring the MSIL1-4 genes (Figure 1B). Notably, a neighboring sub-clade contains two genes, RBGD2 and RGBD4 (Figure 1B), that encode for heat-inducible RBPs implicated in the response to heat stress in Arabidopsis (Zhu et al., 2022). MSIL1-4 proteins share a common domain architecture consisting of two N-terminal RNA recognition motifs (RRM1 and RRM2) that are followed by a poorly conserved, intrinsically unstructured, carboxy-terminal region (Figure 1A and Figure1-figure supplement 1A). The survey of the gene expression ARAPORT11 databases (https://araport.org/; http://fgcz-pep2pro.uzh.ch) revealed that the MSIL1-4 genes are widely expressed in Arabidopsis, with MSIL1 showing the lowest levels of expression (Figure 1C). Database searches further indicated that MSIL orthologs are widely distributed in land plants, from bryophytes to angiosperms (Figure1-figure supplement 1B). To gain a mechanistic understanding of MSIL function in Arabidopsis, we determined the subcellular localization of these proteins by performing fluorescence microscopy on GFP-tagged versions of MSIL2 and MSIL4 (MSIL2G and MSIL4G). Both proteins showed a diffuse cytoplasmic distribution in root cells of stable Arabidopsis transgenic plants that contrasts with the nucleus and cytoplasmic distributions of the free GFP protein control (Figure 1D).

Musashi-like MSIL2/4 proteins redundantly control development in Arabidopsis.

(A) Schematic representation of the Musashi-like (MSIL) protein family in Arabidopsis thaliana. The invariant RNP1 and RNP2 motifs within the conserved RRM domains are indicated. Numbers refer to amino acid identities between the Mouse Musashi/MSI RRM1/2 domains and the corresponding domains in Arabidopsis MSIL homologs. (B) Evolutionary relationships between MSILs and related RNA-binding proteins. The scale bar indicates the rate of evolutionary change expressed as number of amino acid substitutions per site. (C) RNA-seq expression map of MSIL genes extracted from ARAPORT11. Shading is a log2 scale of transcripts per million (TPM). (D) General overview of the roots of 6-days-old wild type Arabidopsis seedlings (Col-0) or transgenic seedlings expressing either a free GFP protein or GFP-tagged versions of MSIL2/MSIL2G and MSIL4/MSIL4G. Scale bar, 10 μm. (E) Photographs of representative rosettes of Col-0, msil2/4, msil2/4-MSIL2F1 and msil2/4-MSIL4F1 plants. (F) Photographs of representative inflorescence stems of Col-0, msil2/4, msil2/4-MSIL2F1 and msil2/4-MSIL4F1 plants. Abbreviations: Mus musculus Musashi2 (MmMsi2); human heterogeneous nuclear ribonucleoprotein A1 (hnRNPA1).

To further investigate the function of the MSILs in Arabidopsis, we raised specific antibodies against non-conserved regions of MSIL1-4 and characterized T-DNA insertion msil mutant lines that lack full-length MSIL mRNA and MSIL proteins as judged by RT-PCR and western blot experiments, respectively (Figure1-figure supplement 2A-C). However, none of the four characterized single msil mutants showed any obvious phenotype (Figure1-figure supplement 3A and data not shown). To assess the potential functional redundancy between MSIL genes, we crossed the single msil mutants together and characterized the six possible double mutant combinations (Figure1-figure supplement 3A). Only the msil2-1/msil4-1 (msil2/4) double mutant showed discernable developmental abnormal phenotypes, including enlarged and curled leaves, early leaf senescence and a pendant inflorescence stem (Figure 1E and F and Figure1-figure supplement 3A-C). To assess whether MSIL1 and MSIL3 also contribute to Arabidopsis development in the msil2/4 double mutant background, we generated a msil1-1/msil2-1/msil3-1/msil4-1 (msil1/2/3/4) quadruple mutant that shows no particular or aggravated developmental phenotypes with respect to the msil2/4 double mutant (Figure1-figure supplement 3A-C). To test the functional redundancy of MSIL2 and MSIL4 proteins in vivo, we transformed the msil2/4 double mutant with constructs coding either for a Flag/HA-tagged version of MSIL2 (MSIL2F) or MSIL4 (MSIL4F), and selected two independent transgenic plants expressing similar and near physiological levels of the tagged protein, as evaluated by western blot using our home-made antibodies (Figure1-Figure supplement 3D). The ectopic expression of either MSIL2F or MSIL4F proteins was able to rescue all the phenotypes associated with the MSIL2/4 defect (Figure 1E and F and Figure1-Figure supplement 3E). This confirms that MSIL2F and MSIL4F are functional proteins and more importantly that MSIL2 and MSIL4 redundantly control various aspects of development in Arabidopsis.

RRM-dependent RNA-binding activity is essential for MSIL2/4 functions in vivo

Animal MSIs interact with target transcripts through their RRM domains and mutations in these domains have been shown to impair MSI activity in vivo (Chavali et al., 2017; Nguyen et al., 2020). MSILs are very similar to metazoan MSIs over their RRM sequences (Figure 1A and Figure 2-figure supplement 1A), a closeness of sequences also observed when homology models of the MSIL4 RRM motifs were generated using SWISS-MODEL (Biasani et al., 2014) and AlphaFold2 (Jumper et al. 2021) servers (Fig. 2A and Figure 2-figure supplement 1B). Homology-based structure prediction proposed that these domains adopt the characteristic RRM fold with a four-stranded β-sheet structure bearing the conserved phenylalanine residues involved in the specific recognition of RNA bases (Daubner et al., 2013) (Figure 2A and Figure 2-figure supplement 1A). This supports the idea that MSILs have retained the ability to interact with RNA, a notion further supported by the identification of MSIL2/4 proteins in the experimentally determined Arabidopsis mRNA-binding proteome (Reichel et al., 2016). To examine the requirement for the interaction of MSIL with RNA in developmental control, we first generated an RNA-binding mutant form of MSIL4, MSIL4RRM, by introducing phenylalanine to aspartate (F→D) mutations in both RRM domains (Figure 2B). Protein-RNA interaction assays using RNA homopolymers immobilized on agarose beads confirmed that the MSIL4, but not MSIL4RRM, exhibited intrinsic RNA binding activity in vitro (Figure 2C). To assess the impact of these mutations in vivo, the MSIL4G and MSIL4GRRM constructs were expressed under the control of the MSIL4 endogenous promoter into the msil2/4 mutant, and two independent transformants expressing near physiological levels of the WT (MSIL4G-3 and MSIL4G-8), and mutant (MSIL4GRRM-3 and MSIL4GRRM-10) proteins were selected for further complementation analysis (Figure 2-figure supplement 2A). MSIL4G, but not MSIL4GRRM proteins were able to rescue the developmental defects incurred by MSIL2/4 mutations (Figure 2D and Figure 2-figure supplement 2B-C), indicating that these proteins are likely to exert their functions by interacting with RNA targets in vivo.

RRM-dependent RNA-binding activity is essential for MSIL function in planta.

(A) Experimentally determined structures of the RRM1 domain of MmMSI2 (PDB ID: 6C8U) and the RRM2 domain of MmMSI1 (PDB ID: 5x3y), and models of the RRM domains of MSIL4 generated using the homology-modeling server SWISS-MODEL. The α-helices and β-sheets of RRM domain as well as the phenylalanine residues that contact RNA bases are highlighted. (B) Schematic representation of the mutations introduced in the RRM domains of MSIL4. (C) Binding assays of MSIL4 RRM domains on ssRNA homopolymers in vitro. Single-stranded polyA (pA), polyG (pG), polyC (pC) or polyU (pU) RNAs were subjected to binding with the His-tagged recombinant RRM domains from MSIL4 (upper part) or its mutated version (MSIL4RRM, lower part). An anti-His antibody was used for detection. In. represents the input fraction. (D) RRM-dependent RNA-binding activity is essential for MSIL2/4 function in stem. Photographs of representative inflorescence stems of Col-0 and msil2/4 mutants complemented or not with a WT version of MSIL4 (MSIL4G-8 and 3) or an rrm mutant (MSIL4GRRM-3 and 10).

MSIL2/4 protein interactomes are enriched in proteins involved in 3’-UTR binding and translational regulation

Animal MSIs modulate mRNA expression mostly at a translational level through binding to conserved motifs in the 3’-UTR of the target mRNA and further interactions with various mRNA binding partners (Fox et al., 2015; Weill et al., 2017; Cragle et al., 2019). To further characterize the components of the MSIL2/4 network in Arabidopsis, we performed affinity purification coupled to LC-MS/MS, as described in Scheer et al. (Scheer et al., 2021), using the complemented MSIL2F1 and MSIL4F1 lines. Consistent with their functional redundancy, MSIL2F1 and MSIL4F1 exhibit a similar protein interaction network, with 27 proteins significantly enriched in both MS analyses (Figure 3A and B, and Supplementary file 1). Examination of the PANTHER database (http://www.pantherdb.org) identified that MSIL2F1/4F1 are part of a protein-protein interaction network that was principally enriched in 4 categories of GO molecular functions, namely poly(A) binding protein (GO:0008143), mRNA 3-UTR binding (GO:0003730), single-stranded RNA binding (GO:0003727), and translation initiation factor activity (GO:0003743), highlighting the relationship between MSIL2/4 and the mRNA binding and regulatory proteome (Figure 3C).

MSIL2/4 protein interactomes are enriched in proteins involved in 3’-UTR binding and translation regulation.

(A) The semi-volcano plots show the enrichment of proteins co-purified with MSIL2F1 and MSIL4F1 as compared with control IPs. Y- and X-axis display adjusted p-values and fold changes, respectively. The dashed lines indicate the threshold above which proteins are significantly enriched/depleted (fold change > 2; adjP < 0.05). (B) Venn diagram showing the overlap between the MSIL2 and the MSIL4 interactomes. (C) PANTHER-based classification of the GO molecular function that are overrepresented among the MSIL-interacting proteins. (D) Confocal monitoring of the colocalization of MSIL2G, MSIL4G and PAB2R in root tips of 8-d-old seedlings were monitored after 30 min of exposure to 38°C (heat stress) or 20°C for control treatment. (E) Cycloheximide treatment inhibits the formation of MSILG and PAB2R cytosolic foci. The transgenic plants expressing the stress granule marker PAB2R was used as a control. Root of 7-day-old seedlings expressing either MSIL4G, MSIL2G, or PAB2R were monitored after 1 hour of exposure to 38°C in the presence of DMSO control treatment (DMSO) or in the presence of cycloheximide inhibitor (+CHX). Scale bar, 10 μm.

Many of the proteins present in the MSIL2/4 interactome, such as PABs, PUMs, UBP1 are involved in translation regulation and are targeted to stress granule (SGs) components, a membraneless cytoplasmic organelle that formed consecutive to the global inhibition of translation caused by heat-stress (Chantarachot and Bailey-Serres, 2018; Scutenaire et al., 2018). To assess the potential functional link between MSIL and those proteins, we crossed the MSIL2G/4G plants with a functional PAB2-RFP/PAB2R line (Scutenaire et al., 2018), and assessed the localization of the fusion proteins before and after heat-stress in both parent and crossed lines. Fluorescence analysis revealed that MSIL2G/4G, like PAB2R, show a dispersed cytoplasmic signal under normal condition, and are recruited to cytoplasmic granules that largely overlap with PAB2R-containing SG upon heat stress as judged by the yellow-orange appearance in the merged figure (Figure 3D and Figure 3-figure supplement 1), and by an independent analysis using the Colocalization Threshold plugin from ImageJ software (data not shown). Treatment with cycloheximide (CHX), a translation inhibitor that prevent SG assembly by trapping mRNAs at polysomes (Mahboubi and Stochaj, 2017), confirmed that those foci correspond to bonafide SGs (Figure 3E). These results were consistent with the RNA-binding protein-based signature and the nature of MSIL interactomes, further supporting a functional relationship between MSILs and the translation machinery in Arabidopsis.

MSIL2/4 proteins regulate the molecular architecture of SCW in Arabidopsis fibers

Pendant stem inflorescence is a common phenotypic trait of Arabidopsis mutants defective in SCW biosynthesis, that results from a lack of mechanical support and rigidity (Zhang et al., 2018; Mitsuda et al., 2007). Given the current lack of knowledge of the role of post-transcriptional mechanisms in the control of SCW biosynthesis in plants, we henceforth focused our analysis on the pendant stem phenotype of the msil2/4 double mutant. Survey of the Arabidopsis inflorescence stem tissue-specific transcriptome database (Shi et al., 2021) indicated that the MSIL2/4 genes are highly expressed in the SCW-forming interfascicular fiber (if) and xylem vessel (xy) cells (Figure 4A). Toluidine blue O staining and microscopic analysis of Col-0 and msil2/4 mutant stem cross-sections revealed no major changes in the architecture of both the interfascicular and xylary fibers in the mutant stem. Notably, xylem cells deformation or collapse, which are a characteristic of mutants deficient in SCW synthesis, were not observed in the msil2/4 mutant (Figure 4B). However, the examination of the cell wall thickness in cross sections of stems revealed that the interfascicular fiber cells display a significantly thinner SCW in msil2/4 mutant (Figure 4C). UV autofluorescence and phloroglucinol staining of the Col-0, msil2/4, msil2/4+MSIL2F1 and msil2/4+MSIL4F1 stem sections revealed a reduced lignin signal in the interfascicular fibers of the msil2/4 mutant compared to Col-0 or complemented plants (Figure 4D and Figure 4-figure supplement 1A). In agreement with these observations, a significant decrease of 30% in AcBr lignin content was observed in the msil2/4 mutant compared to the Col-0 control plants (Figure 4E). Despite a tendency to decrease, the difference of lignin content was not significant in the complemented lines msil2/4+MSIL2F1 and msil2/4+MSIL4F1 compared to Col-0 (Figure 4E). To further investigate potential changes in the lignin structure and composition in the Col-0 and msil2/4 mutant, we evaluated the impact of MSIL2/4 deficiency on lignin composition by thioacidolysis, a degradation method allowing the determination of the relative amounts of H, G and S lignin units linked by beta-O-4 linkages. Total monomer yield did not significantly differ between the Col-0 and the msil2/4 mutant (Figure 4-figure supplement 1B), suggesting that the structure of the lignin polymer is not significantly affected in the mutant. Further examination of β-glucan composition using the fluorescent dye calcofluor white (Hert and Schnepf, 1980) revealed an increased staining in the interfascicular fiber cells of msil2/4 mutant, supporting changes in the composition/accessibility of cellulose (Figure 4F and Figure 4-figure supplement 1C).

MSIL proteins regulate the molecular architecture of SCW in Arabidopsis fibers.

(A) Expression profiles of the MSIL2 and MSIL4 genes in the inflorescence stem tissues as retrieved from the Arabidopsis inflorescence stem tissue-specific transcriptome database40 (https://arabidopsis-stem.cos.uni-heidelberg.de/). The results of two replicates are shown. F, fibers; X, xylem vessels; Cx Cambium (xylem side); Cp, Cambium (Phloem side); P, Phloem; S, Starch sheath; E, Epidermis. (B) Cross-sections of Col-0 and msil2/4 stems showing vascular bundle (xy) and interfascicular fiber (if) cells stained with Toluidine blue. Representative fiber and xylary areas analyzed in the panel 4C are outlined. (C) Measurements of the SCW thickness of xylary fiber and interfascicular fiber cells. Values are means (n>400)± SEM. Data were analyzed by unpaired Student’s test. Asterisks indicate significant differences relative to Col-0; ****p<0.0001. (D) Top: lignin autofluorescence under ultraviolet (UV) using confocal microscopy in Col-0, msil2/4 mutant, and complemented msil2/4-MSIL2F1 and msil2/4-MSIL4F1 plant stem sections. xy: xylem fibers; if: interfascicular fibers. Bottom: Phloroglucinol-HCl staining of Col-0, msil2/4, and complemented msil2/4-MSIL2F1 and msil2/4-MSIL4F1 plant stem sections. Scale bar, 100 μm. (E) Boxplots represent acetyl bromide lignin content, expressed as % of dry weight. Lignin content was measured in the bottom section of mature stems (3 biological replicates per line). Significant differences between Col-0, msil2/4+MSIL2F1, msil2/4+MSIL4F1, and msil2/4 plant lines are indicated by asterisks*, P-value< 0.01 according to Student’s t test (n = 6 to 10). (F) Confocal microscopy of Calcofluor-white staining of cross sections of Col-0 and msil2/4 mutant inflorescence stems. Scale bar, 100 μm. (G) Boxplots represent the glucose yield after incubation of Col-0 and msil2/4 lignocellulosic material with cellulases either without pre-treatment (-), or after pre-treatment (+) with sodium hydroxide (NaOH).

To extend this analysis, we then subjected purified Col-0 and msil2/4 SCW materials to Fourier-Transform Infrared (FTIR) spectroscopy (Alonso-Simón et al., 2011). PLS-DA analyses performed on FT-IR absorbance data clearly separated msil2/4 mutant lines from Col-0 plants, using the first two components (explaining 74% and 10% of total variability, respectively) (Figure 4-figure supplement 1D, left panel). The large-scale chemotyping of CW composition in stem by FTIR spectroscopy suggested that not only lignin deposition was impaired in msil2/4 mutants but also polysaccharides composition (Figure 4-figure supplement 1D, right panel; Supplementary file 2). The lowest absorbance values in msil2/4 lines were observed in the region 1800-1500cm-1, mostly associated to the absorbance of lignin related compounds. On the contrary, a higher absorbance was detected in msil2/4 samples within 1050-900 cm-1 region, mostly associated to polysaccharides compounds (Alonso-Simón et al., 2011). Because the optimal interaction between the SCW lignin and polysaccharide polymers is believed to contribute to SCW recalcitrance, we used saccharification as a proxy to compare cell wall sugars accessibility between Col-0, msil2/4 and complemented mutant lines (Figure 4G). A higher amount of reducing sugars from native cell wall (without pre-treatment/-Tr) was released using msil2/4 samples compared to Col-0 samples. After cell wall loosening using alkali pre-treatment/+Tr, an even more important amounts of sugars (+5%) was released from double mutants CW enzymatic digestion (Figure 4G). Together, our data indicate that MSIL2/4 regulate the molecular architecture of SCW in the interfascicular fibers, contributing significantly to the setting of biomass recalcitrance in Arabidopsis.

The accumulation and activity of the glucuronoxylan decoration machinery are mis-regulated in msil2/4 mutant

To elucidate the impact of the MSIL2/4 on the expression of SCW-related biosynthesis genes in the Arabidopsis inflorescence stem, we performed mRNA sequencing (mRNA-seq) on both wild-type and msil2/4 mutant backgrounds. In total, we found 234 genes to be significantly differentially regulated (DEG) in the msil2/4 mutant background (with stringent thresholds having a logarithm of fold change >1.5 or <-1.5 and false discovery rate (FDR ≤ 0.05), including 156 up-regulated and 78 down-regulated mRNAs (Figure 5-figure supplement 1A, and Supplementary file 3). Interestingly, genes encoding regulatory or enzymatic components of the SCW biosynthesis pathway were absent from the DEG list, and their expression was not significant changed in msil2/4 mutant (Figure 5-figure supplement 1A). Gene Ontology analysis of DEGs revealed significantly enriched signaling pathways (FDR ≤0.05) involved in plant defense or responses to biotic and abiotic stresses (Figure 5-figure supplement 1B), encoding products that are preferentially targeted to the extracellular region and the cell wall (Figure 5-figure supplement 1C).

The low representation of SCW biosynthesis genes in the DEG list suggests that MSIL2/4 could regulate the SCW formation at a post-transcriptional/translational level, a hallmark of MSI function in animal models (Fox et al., 2015; Kudinov et al., 2017). To investigate proteome-level changes that could potentially occur upon MSIL2/4 mutations, we performed a comparative proteomic analysis of Col-0 and msil2/4 inflorescence stems using MS-based quantitative proteomics, and represented ≈4300 reliably identified and quantified proteins in a volcano plot according to their statistical p-value and their relative difference of abundance. This analysis revealed 267 proteins with significant abundances in the msil2/4 mutant versus Col-0 inflorescence stems (fold change ≥2, Benjamini-Hochberg FDR < 1%), including 156 up-regulated, and 111 down-regulated proteins (Figure 5A and Supplementary file 4). While no specific GO terms were enriched in the set of down-regulated proteins, the inspection of the up-regulated proteins revealed a strong enrichment in enzymes involved in the glucuronoxylan biosynthetic pathway (Figure 5A and B, and Supplementary file 4). Notably, the upregulated glucuronoxylan-related proteins included glycosyltransferases required for the xylan backbone synthesis (IRX9/10/14), as well as enzymes involved in its substitution (ESK1, GUX1, and GXM3) (Scheller and Ulvskov, 2010) (Figure 5A-C). Visual inspection of the RNA-seq aligned reads of the up-regulated glucuronoxylan-related genes using the Integrative Genomics Viewer (IGV) and quantitative RT-PCR analysis of the glucuronoxylan-related decoration genes confirmed that the observed proteomic changes in msil2/4 were not associated with significant variations in mRNA levels (Figure 5-figure supplement 2A-B).

The accumulation and activity of the glucuronoxylan decoration machinery are altered in msil2/4 mutant.

(A) MS-based quantitative comparison of Col-0 and msil2/4 inflorescence stem proteomes. Volcano plot displaying the differential abundance of proteins in Col-0 and msil2/4 cells analyzed by MS-based label-free quantitative proteomics. The volcano plot represents the - log10 (limma p-value) on y axis plotted against the log2 (Fold Change msil2/4 vs Col-0) on x axis for each quantified protein. Green and red dots represent proteins found significantly enriched respectively in msil2/4 and Col-0 Arabidopsis stems (log2(Fold Change) ≥ 1 and - log10(p-value) ≥ 2.29, leading to a Benjamini-Hochberg FDR = 1.01 %). The up-regulated proteins involved the glucuronoxylan biosynthetic pathways are indicated. (B) Gene Ontology (GO) analysis of the proteins that are significantly up-regulated in the msil2/4 mutant. The gene ontology analysis of DEGs was performed using ShinyGO v0.76 software. Lollipop diagrams provide information about GO fold enrichment, significance (FDR in log10), and number of genes in each pathway. (C) Schematic model of glucuronoxylan substitution patterns in Arabidopsis and the enzymes involved. IRX9/10/14, glycosyltransferases involved in the synthesis of xylan (Xy) backbone. ESK1, eskimo1. GUX1, glucuronic acid substitution of xylan1; GXM3, glucuronoxylan methyltransferase3 are involved in the glucuronoxylan decoration. (D) Neutral monosaccharide composition of alcohol insoluble residue (AIR) extracted from inflorescence stems of wild-type and msil2/4 mutant plants that have been pre-hydrolyzed or not with acid. Rha, rhamnose; Fuc, fucose; Ara, arabinose; Xyl, xylose; Man, mannose; Gal, galactose; Glc, glucose. (E) Left: Immunofluorescence labeling of transverse sections of Col-0 and msil2/4 stems with the LM10 and LM11 antixylan antibodies. Representative fiber and xylary regions analyzed by immunofluorescence are outlined. Xy, xylem fibers; if, interfascicular fibers. Right: Quantification of the fluorescence was done using ImageJ software and processed according to Supplementary file 5. Significant differences are indicated by asterisks*, P-value< 0.01 according to Student’s t test (n = 5 to 12)

The outcomes of our histological and proteomic analyses suggest potential variations in polysaccharide content and/or composition in the SCW of msil2/4 mutant. To address this point, we carried out monosaccharide composition analysis on either unpretreated (-Tr) or acid-pretreated (+Tr) WT and msil2/4 stem cell walls. The levels of glucose (Glc) and xylose (Xyl), which are the building blocks of cellulose and xylan polysaccharides in the SCW were not substantially affected in the msil2/4 mutant (Figure 5D). Similarly, no major changes were observed in the levels of the primary cell wall-related monosaccharides, galactose, rhamnose, arabinose, mannose, fucose (Figure 5D), indicating that the MSIL2/4 mutations did not significantly altered cellulose or glucuronoxylan levels in the SCW.

To assess possible changes in glucuronoxylan decoration in msil2/4 mutant, we performed cytochemical analysis on Col-0 and msil2/4 stem sections using two antixylan antibodies (LM10 and LM11), whose binding specificities depend on the degree of glucuronoxylan substitution (McCartney et al., 2006). LM11, which binds with similar affinity both unsubstituted and substituted glucuronoxylan, shows a similar signal in the interfascicular and xylary fibers of Col-0 and msil2/4 mutant stems, confirming that the glucuronoxylan content was not significantly modified in msil2/4 compared to Col-0 (Figure 5E). In contrast, LM10, which binds less efficiently the substituted glucuronoxylan (McCartney et al., 2006), shows a significant signal decrease in the interfascicular fibers of msil2/4 mutant (Figure 5E and Supplementary file 5). Together, our data indicate that the MSIL2/4 mutations impact the accumulation and activity of the glucuronoxylan decoration machinery in the interfascicular fiber cells of Arabidopsis stem.

MSILs restrain the degree of 4-O-methylation of glucuronoxylan in Arabidopsis

In Arabidopsis, the glucuronoxylan backbone is decorated predominantly by acetyl, GlcA, and MeGlcA groups, that are deposited in a regular and controlled manner by a specific enzymatic machinery (Scheller and Ulvskov, 2010; Grantham et al., 2017; Wierzbicki et al., 2019). Interestingly, the ESK1, GUX2, and GXM3 enzymes, which are involved in the deposition of the acetyl and GlcA groups, as well as the methylation of GlcA group, respectively, are over-accumulated in the msil2/4 stem proteomics, providing a possible explanation for the modification in glucuronoxylan decoration observed in the interfascicular fiber cells of the msil2/4 mutant. To investigate further the variations in glucuronoxylan decoration incurred by the msil2/4 mutation, we analyzed the patterns of xylanase-released SCW oligosaccharide using matrix-assisted laser desorption ionization-time-of-flight mass spectrometry (MALDI-TOF) (Ropartz et al., 2011). Consistent with previous observations (Mortimer et al., 2010; Urbanowicz et al., 2012), MALDI-TOF analysis of Col-0 glucurononoxylan oligosaccharides showed the release of xylo-oligosaccharide pairs, evenly spaced by their degrees of polymerization and acetylation, bearing a GlcA (m/z 1705.5/1747.5) or methylated GlcA (m/z 1719.5/1761.5) substitution group (Figure 6A). Interestingly, the level of GlcA branched xylo-oligomers was strongly decreased in the msil2/4 double mutant, and this whatever the complexity of the oligomers considered (Figure 6A and Figure 6-figure supplement 1A). Importantly, the molecular phenotypes observed in the msil2/4 mutant were rescued in both msil2/4+MSIL2F1 and msil2/4+MSIL4F1 complemented lines, confirming the specificity of the mutation (Figure 6A). In agreement with the MALDI–TOF outcome, quantification of the acidic sugar levels in unpretreated (-Tr), or acid pretreated (+Tr) WT and msil2/4 stem samples confirmed that the release of GlcA moiety was strongly reduced in msil2/4 double mutant, as opposed to the galacturonic acid/GalA group located on pectins (Figure 6B).

MSILs restrain the degree of 4-O-methylation of glucuroxylan in Arabidopsis inflorescence stem.

(A) MALDI-TOF mass spectra of xylooligosaccharides generated by xylanase digestion of xylan from Col-0, msil2/4, and complemented msil2/4-MSIL2F1 and msil2/4-MSIL4F1 inflorescence stem materials. The ions at m/z 1705/1747 and 1719/1761 correspond to acetylated xylo-decapolysaccharides bearing a GlcA residu (GlcA-Xyl10-Ac4//GlcA-Xyl10-Ac5) or a methylated GlcA residue (MeGlcA-Xyl10-Ac4//MeGlca-Xyl10-Ac5). (B) Acidic monosaccharide composition of alcohol insoluble residue (AIR) extracted from inflorescence stems of wild-type and msil2/4 mutant plants that have been pre-hydrolyzed or not with acid. GalA, galacturonic acid; GlcA, glucuronic acid. (C) Relative changes in unmethylated/methylated GlcA decapolysaccharide ratio in Col-0 (black) and msil2/4 mutant (red) as controlled by the addition of external spike-in control corresponding to a pentaacetyl-chitopentaose. (D) GFP-based RNA-IP assays. Top panel (WB): western blots performed using antiGFP or antiUGPase antibodies on protein fractions from inputs or antiGFP immunoprecipitates from WT and the msil2/4 mutant plants expressing the WT (MSIL4G-3) or RRM mutant (MSIL4GRRM-3) MSIL4-GFP fusions. Bottom panel (RT) : Corresponding RT-PCR using GXM1 or GXM3 specific primers on RNA fractions.

It has been previously proposed that the intermediate level of glucuronoxylan methylation observed in Arabidopsis is due to an intrinsic rate-limiting activity of GXM enzymes in vivo (Yuan et al., 2014). We noticed that in addition to GXM3, a second member of the GXM family, GXM1 (Yuan et al., 2014), tends also to be over-expressed in msil2/4 stem although at a fold change of 1.77 slightly below the fixed significance threshold in the proteomic analysis (Figure 5A and Supplementary file 4). This observations make us wonder whether the apparent decrease of GlcA branched xylooligosaccharides in msil2/4 could be due to an overmethylation of the GlcA substituent. To address this question, we performed the xylanase digestion and MALDI-TOF analysis in presence of an added spike-in oligosaccharide control (Pentaacetyl-chitopentaose) for normalization of the MALDI_TOF data (Figure 6-figure supplement 1A). Upon normalization, the MALDI-TOF data indicated that the decrease of the peaks corresponding to the GlcA-branched xylooligosaccharides was correlated with a increase in corresponding meGlcA branched xylooligosaccharides, (Figure 6C and Figure 6-figure supplement 1B, Supplementary file 6), suggesting that the MSIL2/4 mutations unleash the activity of the GXM1/3 in stem. RNA immunoprecipitation assays further confirmed that the GXM3 and GXM1 mRNAs are efficiently pulled-down from MSIL4G-3, but not MSIL4GRRM-3 plant extracts (Figure 6D), suggesting that the MSIL2/4 interaction network controls the 4-O-methylation of glucuronoxylan by repressing the activity of the GXM1/3 mRNAs in stem.

Discussion

RNA-binding proteins (RBPs) are essential components of the gene regulation machinery that govern the fate and expression of cellular RNA at post-transcriptional levels, including processing, splicing, base modification, translation and degradation (Singh et al., 2015). Sequence-based bioinformatic, reverse and forward genetic, and affinity-based proteomic analyses have converged to show that plants, like animals and fungi, harbor a large diversity of RBPs whose functions in gene regulation and plant development are ill-defined (Reichel et al., 2016). In this study, we have identified an RRM motif-containing RBP family, hereafter named as MSIL, whose members share sequence, structural and functional similarities with the animal translational regulator Musashi/MSI. We found that two of the four MSIL members, MSIL2 and MSIL4, function redundantly to control various aspects of Arabidopsis development, including leaf senescence, morphology of rosette leaves, and the rigidity of the inflorescence stem. Our data suggest that the pleiotropy associated with the MSIL2/4 mutation likely results from the multifunctionality of these proteins in Arabidopsis development, a situation similar to the complex phenotypic traits previously associated to MSI mutations in animals (Fox et al., 2015; Kudinov et al., 2017; Nguyen et al., 2020). Although the molecular bases underlying the msil2/4 phenotypic variation remain unclear, the presence of MSILs in the high confidence Arabidopsis mRNA-binding proteome (Reichel et al., 2016), suggests that these proteins are likely to exert control over development by targeting a complex repertoire of mRNAs. Regardless of the mechanism, our data support a cell/tissue-context specific role for MSIL and provide a resource for further studies on the mechanisms of MSIL regulation in plants.

Our data also provide evidence that the RRM domains of the MSIL2/4 proteins exhibit RNA-binding ability and are essential for the activity of these proteins in vivo, an observation that is corroborated by the presence of the MSIL2/4 proteins in an experimentally determined Arabidopsis mRNA-binding proteome (Reichel et al., 2016). Further supporting a role for MSIL2/4 in mRNA-related transactions, affinity purification-mass spectrometry analysis revealed that these proteins share a common network of protein-protein interactions with well-known mRNA-binding proteins, including the PAB2/4/8, RBP45/47, and the translation factor EIF4G. Notably, the animal polyA-binding protein/PABP (homolog of the plant PABs) was previously shown to be a functional partner of MSI that is essential for it activity in translation repression (Fox et al., 2015; Cragle et al. 2019). Interestingly, the MSIL2/4-associated proteins also include the phylogenetically related RBGD2/4 proteins, two constitutively expressed RBPs that are involved in heat resistance in Arabidopsis (Zhu et al., 2022). The comparison of the MSIL2/4 and RBGD2/4 interactomes indicates only a partial overlap, suggesting a certain level of subfunctionalization. In accordance with this observation, the MSILs and RBGDs clearly diversify in function, since the rbgd2/4 double mutant exhibits no developmental phenotypes under normal growth conditions.

A penetrant developmental phenotype associated with the MSIL2/4 mutations is a loss in stem rigidity that we could trace back to a defect in SCW formation, an important reservoir of fixed carbon and a renewable resource of major environmental and economic importance (Meents et al., 2018). Yet, with the exception of few microRNAs involved in the control of lignin polymer synthesis (Zhang et al., 2018; Wang et al., 2014; Zhao et al., 2015), no information about RBPs-mediated posttranscriptional regulation of SCW biosynthetic genes was previously reported. In this study, we show that MSIL2 and MSIL4, function redundantly to promote SCW formation in the interfascicular fibers and that their ability to bind RNA is essential to fulfill this function. Our conclusions are supported by independent approaches which highlight changes of SCW thickness, lignin content, and xylan decoration profile in the interfascicular fiber cells of the msil2/4 mutant. The fiber-specific effect of the MSIL2/4 mutations on SCW formation, is further supported by our evidence showing that the msil2/4 mutant grows normally and lacks the xylem collapses usually observed in more pleiotropic SCW deficient mutants. Previous studies have shown that specific master transcriptional regulators, such as NST1 and NST3, control cell-lineage specific formation of SCW formation in Arabidopsis (Taylor-Teeples et al., 2015; Mitsuda et al., 2007). Our study extends this observation by showing that RBP-mediated post-transcriptional mechanisms are also involved in cell-type specific control of SCW formation in Arabidopsis. Notably, the presumed role of MSIL2/4 in SCW formation in the interfascicular fiber cells contrasts with the rather constitutive expression of the corresponding MSIL2/4 genes, suggesting as previously discussed, that MSIL2/4 activities depend on cell-context. This is in accordance with the reported cell-type specific activities of the ubiquitous MSI2 in cell renewal, cell differentiation and cancer control in animals (Fox et al., 2015; Kudinov et al., 2017). Understanding how the activity of the MSIL2/4 genes is controlled by the cellular context will be essential from a basic point of view, providing a unique model for studying RBP regulation in specific cell type in plants.

From the set of genes found to be misregulated in the msil2/4 stem transcriptome, none could be functionally associated with SCW biosynthesis. In contrast, we found that the MSIL2/4 mutation upregulates a battery of plant defense-related genes, known to be involved in systemic acquired resistance (SAR). In this regard, the constitutive activation of defense pathways has been well documented in mutant plants defective in cell wall formation, including the SCW (Ellis et al. 2002; Miedes et al., 2014; Liu et al. 2023), an observation that tends to support the idea that the transcriptome changes incurred by the MSIL2/4 disruption could be indirect. This hypothesis was supported by several lines of evidence: 1) the level of activation of defense genes in msil2/4 is low compared to that usually observed in plants developing SAR or in mutant plants showing an auto-activation of plant immunity (Li et al., 2010; Wu et al., 2012); 2) SAR-dependent activation of defense genes is controlled mostly at a transcriptional level (Hussain et al., 2018); 3) the stem phenotypes displayed by msil2/4 mutant have not been reported, to our knowledge, in the numerous plant-pathogen studies published so far; 4) the top enriched proteins in the stem proteome of msil2/4 are players of the SCW biosynthesis pathway but not plant-defense signaling. That being said, we can’t exclude that the activation of immune-related genes, that include many cell wall-modifying enzymes, additively contributes to the msil2/4 stem phenotype. Further studies will be required to clarify this issue.

Our data reveal that the accumulation of the mRNA of SCW biosynthesis enzymes and regulators are unaffected in the msil2/4 mutant, suggesting that the MSIL2/4 proteins act at a translational level to regulate the SCW formation. In this regard, the stem proteomics gives us a glimpse into the changes in protein abundance that could account for the observed msil2/4 SCW phenotype. Indeed, our MS data reveal that proteins related to the glucuronoxylan biosynthesis machinery, including enzymes involved both in the synthesis (IRX9/10/14) and substitution (GUX1, ESK1, and GXM3) of the xylan backbone, are significantly over-accumulated in the msil2/4 mutant. Knowing that the synthesis of the structurally complex glucuronoxylan polysaccharide requires the coordinated production and action of these enzymes in the Golgi apparatus, our results suggests that MSIL2/4 is exerting a multilevel control on the glucuronoxylan pathway, a mechanism reminiscent of the multi-nodal control displays by MSI2 on the TGF-β and translation pathways in animal cells (Park et al. 2014; Duggimpudi et al., 2018).

From a mechanistic point of view, however, our data raise several intriguing questions. In particular, the extent of glucuronoxylan-related protein abundance changes in msil2/4 is at odds with the minor changes observed to glucuronoxylan composition. In particular, while the proteomic changes can correlate with the observed increase in glucuronoxylan methylation, it is not associated as expected with an increase in global glucuronoxylan content. One possible explanation for this observation could be that the activities of the GUX, ESK, IRXs glucuronoxylan biosynthesis enzymes, in contrast to that of GXM3, are not rate-limiting for glucuronoxylan synthesis in vivo. Alternatively, in light with prior studies that have shown that the steady-state levels of glucuronoxylan and cellulose are correlated within the SCW (Wu et al., 2009; Ye and Zhong, 2022), one could imagine that mechanisms acting downstream of glucuronoxylan de novo synthesis step contribute to fine tune the final content in glucuronoxylan in the SCW. Comparing the levels of glucuronoxylan produced in the Golgi apparatus versus the ones deposited in the SCW in both WT versus msil2/4 plants will be necessary to ascertain this hypothesis. In contrast, the concomitance between the glucuronoxylan hypermethylation and the GXM3 overaccumulation phenotypes observed in msil2/4 clearly support the previous assumption that GXM-mediated methylation is a rate-limiting step in glucuronoxylan that can be overcome by GXM protein overexpression (Yuan et al., 2014). In fine, our data support the idea that transcriptional and post-transcriptional regulation act in a concerted manner to fine tune the GXM3 and glucuronoxylan methylation levels in the stem. We currently posit that MSIL2/4 act in addition to the transcriptional network to adjust the final level of GXM enzymes by regulating the translation of the GXM mRNAs (Fig. 7). Future biochemical and functional studies will be necessary to precisely understand the mechanism by which MSIL2/4 proteins impact gene expression during SCW synthesis in Arabidopsis.

Model of MSIL-dependent control of glucuronoxylan methylation in Arabidopsis and its consequence for SCW architecture.

In the interfascicular fiber cells of Col-0, MSIL2/4 restrain the translation of the glucuronoxylan biosynthesis enzymes, including the rate-limiting GXM3 enzyme. This activity would keep the level of glucuronoxylan methylation at an intermediate level, therefore providing a biochemical environment that favors the interactions between the glucuronoxylan and lignin polymers. In the interfascicular fiber cells of msil2/4 mutant, the translation of the glucuronoxylan biosynthesis machinery, including GXM3, is increased, leading to the deposition of an over-methylated form of glucuronoxylan that would be less prone to interact with lignin and establish SCW formation. In a non exclusive manner, MSIL2/4 could also have a positive role in lignin synthesis, whose defect in msil2/4 would lead to a decrease in lignin content. The model was inspired from Grantham et al., 2017.

Our data indicate that, in addition to the reported change in glucuronoxylan methylation, the MSIL2/4 mutations is also associated with a specific decrease of lignin content in the SCW of the interfascicular fiber cells, whose origin remains unclear. On the basis of the previous observations, we envision two possible mechanisms by which the MSIL2/4 mutations could impact the lignin deposition in the fiber cells. In the first model, we propose that MSIL2/4 positively control the lignin synthesis pathway in the interfascicular cells by promoting the translation of lignin biosynthesis genes (Fig. 7). In this regard, although no GO enrichment could be observed for lignin biosynthesis related genes in the proteomic analysis, two lignin biosynthetic enzymes, CAD4 and PAL2, are significantly down-accumulated in msil2/4 mutant (Supplementary file 4), an observation that could account for the observed decrease in lignin content. However, previous published biochemical and genetic analyses do not support this conclusion, as the single cad4 and pal2 knock-out mutants display no significant alteration of lignin content due to genetic redundancy (Sibout et al., 2005; Rohde et al., 2004). Moreover, thioacidolysis analyses showed that the ratio or degree of polymerization of the different lignin monomers remain unchanged in the msil2/4 mutant, reinforcing the idea that the lignin biosynthesis pathway is functional in this mutant, leaving open the question of the specificity of MSIL2/4 action on the lignin pathway (Fig.7). An alternative explanation for the decrease in lignin phenotype observed in the msil2/4 mutant would be that it results from an indirect cascade effect primarily triggered by the msil2/4-dependent changes in glucuronoxylan methylation. In this regard, it has been proposed that the 4-O-methylation of GlcA substituent results in the xylan being more hydrophobic, possibly impacting its interaction with lignin and the capacity of lignin to assemble in the SCW (Urbanowicz et al., 2012; Busse-Wicher et al., 2013; Kang et al., 2019; Terret and Dupree, 2019) (Fig. 7). However, a 35S promoter-dependent overexpression study of GXMs in Arabidopsis previously failed to report any discernable developmental or molecular phenotypes associated to a glucuronoxylan hypermethylation (Yuan et al., 2014). In the light of these results and despite it is necessary to elucidate deeper the role of MSIL on other SCW genes, our work provides evidence that they are new players in the regulation of SCW and may consequently impact biomass digestibility, with possible implications for improving feedstock for biorefinery.

Methods

Plant material

Arabidopsis thaliana Col-0 and mutants msil1-1 (GABI_462A04), msil2-1 (Salk_066670), msil3-1 (Salk_002477) and msil4-1 (Salk_094167) obtained from the SALK Institute genomic analysis laboratory were used in this study. PAB2-RFP/PAB2R line has been provided by Cécile Bousquet-Antonelli (LGDP, Université Perpignan). The primers used to genotype lines are listed in Supplementary file 7. Double mutants and quadruple mutants were generated by crossing plants. Plants were either grown in soil or cultivated in vitro on plates containing 2.20 g/l synthetic Murashige and Skoog (MS) (Duchefa) medium and 0.5g/l MES, pH 5.7, and 7 g/l agar. To breakdown dormancy, seeds were incubated for 48 h at 6°C in the dark. Germination and development were performed in growth chambers, at 20°C, 60–75% hygrometry with a 16-h light/8-h dark photoperiod (100 μE m −2 s−1 light [fluorescent bulbs with white 6500K spectrum, purchased from Sylvania]). For in vitro culture, the seeds were surface-sterilized before been sown on plates, incubated for 48 h at 6°C in the dark, and placed in a growth cabinet at 20°C with a 16-h-d/8-h-dark cycle and 120 μE m −2 s−1 light (LEDs with white 4500K spectrum, purchased from Vegeled).

Transgenic lines construction

MSIL2F- and MSIL4F-tagged versions were produced from Phusion-generated genomic DNA PCR products (promoter and coding region) using primers TL976(SalI)-TL924(PstI) for MSIL2F and TL974(SalI)-TL975(BamHI) for MSIL4F. PCR products were cloned in the CTL235 binary vector containing a Flag-HA tag in front of a NosT terminator and the hygromycin resistance gene driven by a 35S promoter. To obtain the MSIL2G and MSIL4G constructs used for fluorescence microscopy, the same PCR products were cloned in the CTL579 binary vector in front of a EGFP tag. The obtained plasmids were then used to transform mgl2/4 plants via Agrobacterium transformation. MSIL4-GFP/MSIL4G-tagged version used for complementation and RIP corresponds to the fusion of a genomic PCR product containing MSIL4 promoter (primers TL3527(HindIII)-TL3528(SalI)) fused with a second PCR cDNA fragment (primers TL3529(SalI)-TL3530(BamHI)) cloned initially in pGEM T Easy (Promega). After sequencing the fusion DNA has been cloned in the binary vector CTL579 containing GFP cDNA sequence. The MSIL4G-tagged construct was finally introduced in msil2/4 plants via Agrobacterium transformation. The rrm mutated MSIL4GRRM version was produced as MSIL4G except that the cDNA fragment has been gene-synthetized (Genecust) to modify RNA binding ability by having aspartic instead of phenylalanine at position 9, 49 and 51 in RRM1 domain and in position 111, 151 and 153 in RRM2 domain. Transgenic seeds selection was performed in the presence of 30 μg/l hygromycin and the progenies were screened by PCR.

Protein extraction and immunodetection

Total plant extracts (up to 100 mg) were ground in liquid nitrogen and proteins were extracted according to Hurkman and Tanaka, 1986. Before migration, SDS–PAGE loading buffer was added and Coomassie staining was used to calibrate loadings. Proteins were separated on SDS/ PAGE gels and blotted onto Immobilon-P PVDF membrane (Cat. No. IPVH00010; MerckMillipore). Protein blot analysis was performed using the Immobilon Western Chemiluminescent HRP Substrate (Cat. No. WBKLS0500; MerckMillipore). Specific MSIL antibodies were raised in rabbits against defined epitopes (Epitope MSIL1 (SCDGTSSTFGYNRIPS), MSIL2 (RLQEYFGKYGDLVE), MSIL3 (GYGVKPEVRYSPAVGN) and MSIL4 (TWRSPTPETEGPAPFS) (Eurogentec).

Protein production and purification

The MSIL4 RRM binding domains were amplified using primers TL4128 and TL4129 either from MSIL4G construct for the wild type version or from MSIL4GΔr for the mutated version. PCR products were then cloned in pET41 to obtain His-tagged RRM domains. Tagged proteins were produced by induction in Escherichia coli BL21 cells. Cultures of 200 mL of an ampicillin-resistant (100 mg mL–1) colony were grown at 37°C and induced by 100mM isopropyl-b-D-galactopyranoside in the exponential phase (optical density 0.5 at 500 nm). After induction, bacteria were harvested by centrifugation and pellets were resuspended in 5 ml of binding buffer (His-bind buffer kit (Millipore), proteins were extracted using a Constant cell disrupter system (Constant Systems) with a disruption pressure of 2.35kbar and then purified using His-bind buffer kit (Millipore) following manufacturer’s recommendations.

RNA binding assays

Binding assays were performed using homopolymer RNA conjugated to CNBr-activated Sepharose beads (Sigma C9142). Either poly(A), poly(G), poly(C) or poly (U) from Mercks (respectively P9403, P4903, P4404 and P9528) were used following Sigma’s recommendations. 0.7 to 1 mg of RNA were used for 200 mg sepharose beads. 0.45 mg of purified protein (MSIL4 RRM binding domain or mutated version) in 0.5 ml binding buffer (100 mM NaCl, 10 mM Tris-HCl, pH 7.5, 2.5 mM MgCl2, 2.5 % Triton X-100) were incubated with 5 mg RNA conjugated Sepharose beads for 1 hour. After incubation, beads were washed 5 times for 10 min each in 0.5 ml of binding buffer. Beads were then pelleted at 12,000 rpm, resuspended in SDS sample buffer, boiled for 20 min, and pelleted again. Supernatants were transferred to fresh tubes and loaded on a 12% SDS-PAGE minigel for electrophoresis (Bio-Rad) and protein blot analysis.

RNA immunoprecipitation

RIP was performed as described previously in Merret et al., 2017) using 400 mg of stem powder as starting material.

Expression analysis

Total RNA was isolated using the TRI re-agent (Cat. No. TR-118; Euromedex) according to the manufacturer’s instructions. Genomic DNA was then digested out using the RQ1 RNase-Free DNase (Cat. No. M6101; Promega). cDNAs were obtained from 400 ng of RQ1-treated RNA using 1 μl of GoScript Reverse Transcriptase (Cat. No. A5003; Promega) in a 20-μl final volume reaction using random primers (Cat. No. C1181; Promega) in the presence of 20 units of RNasin (Recombinant Ribonuclease Inhibitor, Cat. No. C2511; Promega) and dNTP mix at a final concentration of 0.5 mM of each dNTP (Cat. No. U1511; Promega). Semi-quantitative RT–PCR amplifications were performed on 1 μl of cDNA in a 12.5-μl reaction volume to start with. The amplification of EF1αtranscripts was used to equilibrate. qRT–PCR was performed on a Light Cycler 480 II machine (Roche Diagnostics) by using the Takyon No ROX SYBR MasterMix blue dTTP kit (Cat. No. UF-NSMT-B0701; Eurogentec). Each amplification reaction was set up in a 10-μl reaction containing each primer at 300 nM and 1 μl of RT template in the case of total RNA with a thermal profile of 95°C for 10 min and 40 amplification cycles of 95°C, 15 s; 60°C, 60 s. Relative transcript accumulation was calculated using the ΔΔCt methodology (Livak and Schmittgen, 2001), using ACTIN2 as internal control. Average ΔΔCt represents three experimental replicates with standard errors. PolyA+ RNAs were purified from RNA obtained by the TRI Reagent extraction, using the PolyATtract mRNA Isolation System III (Cat. No. Z5310; Promega) according to the manufacturer’s instructions. cDNAs were then synthetized according to the above protocol using 50 ng of polyA+ RNA as starting material. Semiquantitative and quantitative RT–PCRs were performed as described above with a 1/25 dilution of cDNAs from polyA+ RNA as template for qRT–PCR. The primers used for RT and qRT-PCR are listed in Supplementary file 7.

RNA Sequencing

Total RNA were extracted from the basal stem section of mature Col-0 and msil2/4 mutant plants using the RNeasy Plant Mini Kit (Qiagen, Cat.#74904) and treated onto the column with the RNase-Free DNase Set (Qiagen, Cat.#79254). Two replicates were performed per plant lines. PolyA purification, library preparation using the Illumina TruSeq stranded mRNA kit and library quality controls were performed by the BioEnvironment Illumina Platform. The mRNA sequencing of Col-0 and msil2/4 mRNAs was done with single-reads, 1x 75pb and 30–35x106 reads per library.

RNASeq analysis

For each library, 30–35x106 reads were obtained with 85–90% of the bases displaying a Q-score ≥ 30, with a mean Q-score of 38 as assessed with Fastqc (http://www.bioinformatics.babraham.ac.uk/projects/fastqc/). Reads were mapped against the TAIR 10 genome using gtf annotation file and default parameters of TopHat2 v2.0.7 (Langmead and Salzberg, 2012). Assembly and transcript quantification were performed with Cufflinks v2.2.1 (Kim et al., 2013). Finally, the low expressed transcripts -less than 1 RPKM (reads per kilobase per million mapped reads) in one of the libraries were filtered out. The differential analysis was conducted by Cuffdiff software, belonging to the Trapnell suite. Foldchange (FC) was determined as the ratio between the normalized read counts between mutant and wild-type.

Fluorescence detection and tissue colorations

GFP detections were performed on in vitro grown one week seedling roots fixed in fixation buffer (50mM Pipes, 5mM EGTA, 5mM MgSO4, pH7 added with 2% para-formaldehyde and 0.2% Triton). After 5 min vacuum infiltration and 10 min agitation at room temperature, roots were washed twice with fixation buffer and stored at 4°C until confocal observation. Confocal imaging was done using an Zeiss Axio Observer Z1 LSM 700. Roots were observed in water and for GFP detection samples were excited at 488 nm and the signal captured at 495–555 nm. For RFP detection, excitation was performed at 555 nm and the signal captures at 600-700 nm. Images were analysed with ImageJ. At least 15 samples per line were analyzed.

Stem coloration were performed from 5 cm length basal stem pieces conserved in ethanol 80%. Stems were rehydrated first in ethanol 50% and later in water and then embedded in agarose 5% before being transversally cut using a vibratome (Leica vt 1000s) to obtain 120 micron thickness sections stored in water. The presence of lignin was determined by staining with phloroglucinol for 30 seconds, giving a red coloration in the presence of lignin cinnamaldehyde groups. Sections were observed using an inverted microscope (Leitz DMRIBE, Leica Micro-systems, Wetzlar, Germany) and images were registered using a CCD camera (Color Collview, Photonic Science, Milham, UK).

For cellulose detection, the brightener Calcofluor White ST [4,4’-bis(anilino-6-bis(2-hydroxyethyl)-amino-s-triazine-ylamino)-2,2’-stilbene disulfonic acid; Cyanamide Co., Bound Brook, N.J] was used at a final concentration of 0.1%. The samples were incubated with calcofluor for 3 min, washed three times with water before being observed. For detection, excitation was performed at 405 nm and the signal captures at 434-496 nm. For auto-fluorescence analysis stem sections were excited at 405 nm and the signal captures at 410-470 nm for blue. Pictures have been analyzed with LAS X (Leica Application Suite X).

Biochemical analyses of SCWs

Arabidopsis stems (5 cm sections at the bottom part of the stems) were harvested from Col0 and msil2/4 double mutant as four independent biological replicates. Soluble extractives were eliminated as described by Ployet et al. (Ployet et al., 2018). Briefly, stems were freeze dried for 48h, ground using a Mixer Mill MM 400 (Retsch), and extracted by hot solvents (successively water, ethanol, ethanol/toluene (1:1 v/v) and acetone) to obtain extractive-free stem residues (ESR). The determination of acetyl bromide (AcBr) lignin was performed on 10mg of ESR as described by Ployet et al. (Ployet et al., 2019). All analyses were done in triplicate. Fourier transform infrared spectroscopy (FT-IR) analysis was performed on 100-200 mg of ESR. Spectra were recorded from ten technical replicates, in the range 400–4000 cm−1 with a 4 cm-1 resolution and 32 scans per spectrum using an attenuated total reflection (ATR) Nicolet IS50 FT-IR spectrometer (Thermo Fisher, Illkirch-Graffenstaden, France) equipped with a deuterated-triglycine sulfate (DTGS) detector. Spectra analyses were performed as described in Dai et al., 2020. Saccharification yield was estimated with or without alkali pretreatment, as described by Van Acker et al., 2016. Reducing sugar concentration was assessed with dinitrosalicylic acid (DNS) reagent using 10µl of the supernatant after 9 h of reaction. Enzyme activity was assessed at 0.25 filter paper units (FPU) ml-1 in our conditions. The Boxplots were generated using ggplot2 R package (R v2022.07.1).

Mass spectrometry (MS)-based proteomic analyses

Proteins from total inflorescence stem extracts of three biological replicates of Col-0 and msil2/4 mutant plants were solubilized in Laemmli buffer and heated for 10 min at 95°C. They were then stacked in the top of a 4-12% NuPAGE gel (Invitrogen), stained with Coomassie blue R-250 (Bio-Rad) before in-gel digestion using modified trypsin (Promega, sequencing grade) as previously described (Casabona et al., 2013). The resulting peptides were analyzed by online nanoliquid chromatography coupled to MS/MS (Ultimate 3000 RSLCnano and Orbitrap Exploris 480, Thermo Fisher Scientific) using a 180-min gradient. For this purpose, the peptides were sampled on a precolumn (300 μm x 5 mm PepMap C18, Thermo Scientific) and separated in a 75 μm x 250 mm C18 column (Aurora Generation 2, 1.6µm, IonOpticks). The MS and MS/MS data were acquired using Xcalibur 4.4 (Thermo Fisher Scientific).

Peptides and proteins were identified by Mascot (version 2.8.0, Matrix Science) through concomitant searches against the Viridiplantae database (from Uniprot, Arabidopsis thaliana (thale-cress) taxonomy, 136447 sequences) and a homemade database containing the sequences of classical contaminant proteins found in proteomic analyses (human keratins, trypsin…, 126 sequences). Trypsin/P was chosen as the enzyme and two missed cleavages were allowed. Precursor and fragment mass error tolerances were set at respectively at 10 and 20 ppm. Peptide modifications allowed during the search were: Carbamidomethyl (C, fixed), Acetyl (Protein N-term, variable) and Oxidation (M, variable). The Proline software (Bouyssié et al., 2020) (version 2.2.0) was used for the compilation, grouping, and filtering of the results (conservation of rank 1 peptides, peptide length ≥ 6 amino acids, false discovery rate of peptide-spectrum-match identifications < 1% (Couté et al., 2020), and minimum of one specific peptide per identified protein group). MS data have been deposited to the ProteomeXchange Consortium via the PRIDE partner repository (Perez-Riverol et al., 2022). Proline was then used to perform a MS1 label-free quantification of the identified protein groups based on razor and specific peptides.

Statistical analysis was performed using the ProStaR software (Wieczorek et al., 2017) based on the quantitative data obtained with the four biological replicates analyzed per condition. Proteins identified in the contaminant database, proteins identified by MS/MS in less than two replicates of one condition, and proteins quantified in less than three replicates of one condition were discarded. After log2 transformation, abundance values were normalized using the variance stabilizing normalization (vsn) method, before missing value imputation (SLSA algorithm for partially observed values in the condition and DetQuantile algorithm for totally absent values in the condition). Statistical testing was conducted with limma, whereby differentially expressed proteins were selected using a log2(Fold Change) cut-off of 1 and a p-value cut-off of 0.00513, allowing to reach a false discovery rate close to 1% according to the Benjamini-Hochberg estimator. Proteins found differentially abundant but identified by MS/MS in less than two replicates, and detected in less than three replicates, in the condition in which they were found to be more abundant were invalidated (p-value = 1).

MALDI-TOF MS analyses of the patterns of xylanase-released SCW oligosaccharide

10 mg of dried alcohol insoluble cell wall residues were incubated with 10U.ml-1 of endo-1,4-D-xylanase (E-XYLNP) from MEGAZYME previously desalted with centrifulgal filters (Amicon Ultra-0.5ml) in 1 ml during 12hrs at 40°C. After boiling during 5 min to inactivate enzyme activity, samples where filtred with 50KD centrifulgal filters (Amicon Ultra-0.5ml). As positive control, 2 mg of wheat AX (MEGAZYMZE) were treated in a similar manner. The samples were then analyzed by matrix-assisted laser desorption ionization-time of flight (MALDI-TOF) MS using DMA/DHB (N,N-dimethylaniline/2,5-dihydroxybenzoic acid) matrix (Ropartz et al., 2011). The samples (1 μL) were deposited and then covered by the matrix (1 μL) on a polished steel MALDI target plate. MALDI measurements were then performed on a rapifleX MALDI-TOF spectrometer (Bruker Daltonics, Bremen, Germany) equipped with a Smartbeam 3D laser (355 nm, 10000 Hz) and controlled using the Flex Control 4.0 software package. The mass spectrometer was operated with positive polarity in reflectron mode. Spectra were acquired in the range of 560–3200 m/z. The detected species correspond to ions in the form of sodium adducts.

Thioacidolysis

This method is adapted from Lapierre et al., 1995. Briefly, 10 mg of alcohol insoluble cell wall residues were incubated in 3ml of dioxane with ethanethiol (10%), BF3 etherate (2.5%) containing 0.1% of heinecosane C21 diluted in CH2Cl2 at 100°C during 4 hours. 3ml of NaHCO3 (0.2 M) were added after cooling and mixed prior the addition of 0.1 mL of HCl (6 M). The tubes were wortexed after addition of 3 mL of dichloromethane and the whole lower organic phase was collected in a new tube before concentration under nitrogen atmosphere to approximately 0.5 ml. Then, 10 μL were trimethylsilylated (TMS) with 100 μL of N,O-bis(trimethylsilyl) trifluoroacetamide and 10 μL of ACS-grade pyridine. The trimethylsilylated samples were injected (1 μL) onto an Agilent 5973 Gas Chromatography–Mass Spectrometry system. Specific ion chromatograms reconstructed at m/z 239, 269 and 299 were used to quantify H, G and S lignin monomers respectively and compared to the internal standard at m/z 57, 71, 85).

Cell wall sugar analysis

Neutral monosaccharide sugar content was determined by gas chromatography after acid hydrolysis and conversion of monomers into alditol acetates as described in Hoebler et al., 1989, and Blakeney et al., 1983. Gas Chromatography was performed on a DB 225 capillary column (J&W Scientific, Folsorn, CA, USA; temperature 205 °C, carrier gas H2). Calibration was made with standard sugar solution and inositol as internal standard.

For acid sugar analysis, the uronic acid content was automatically quantified after hydrolysis of polysaccharides in concentrated sulfuric acid (18 M) containing sodium tetraborate (12.5 mM) or not, followed by m-hydroxydiphenyl colorimetric determination (Blumenkrantz et al., 1973; Thibault, 1979). Galacturonic and Glucuronic acid were used as standard for quantification.

Immunohistochemistry

Samples of Arabidopsis tissues were dehydrated in a successive ethanol series (20%, 40%, 60%, 80%, 95%, and 100%) and embedded in LR White resin (Electron Microscopy Sciences ). Thin sections (1 µm) were placed on Teflon-coated slides (Electron Microscopy Sciences), blocked in phosphate-buffered saline, 2% Tween, and 1% bovine serum albumin for 2 h (PBST-BSA), and labeled overnight (12 h) at 4 °C with primary antibody diluted in PBST-BSA. Sections were washed with PBST and incubated at room temperature for 2 h with a secondary antibody diluted in PBST-BSA. Slides were then washed with deionized water and dried under a stream of dry air. Primary antibodies and dilutions were as follows:, LM10 and LM11 (1:10, v/v; Plant Probes for both). The secondary antibody was a goat anti-rat IgG coupled to the fluorescent dye Alexa Fluor 555 (Molecular Probes) and was used at a 1:100 (v/v) dilution. Observations were carried out using a LeicaDM 6000microscope with a TCS SP8 scanning confocal system using a 40 x apochromatic oil immersion objective to visualized Alexa555 fluorescence (excitation: 561 nm; emission: 565–645 nm). Acquisition settings, including laser power,photomultiplier gain, field of view (X,Y dimension), Z-step, and pixel size of the image, were strictly identical to ensure reliable comparisons between plant material (i.e. the wild type versus mutants).

Data availability

The high-throughput sequencing data generated in this study are accessible at NCBI’s Gene Expression Omnibus (GEO) via GEO Series accession number GSE223732. The mass spectrometry proteomics data have been deposited to the ProteomeXchange Consortium via the PRIDE partner repository with the dataset identifier PXD040207 (proteomic identification of MSIL-interacting partners), and PXD040020 (proteomic analysis of total inflorescence stem extracts).

Supporting information